---
title: "Homework 8"
author: "Sarah Cannon"
date: 'Due: 11/5/2024'
output:
  html_document: default  
  pdf_document: default
---

```{r, message = FALSE, echo = FALSE}
library(DBI)
library(RSQLite)
```

## Question 1 (3 points)

> **As we get farther into this semester, rather than providing you with specific ethics resources, I'm going to start asking you to do some of your own research!  This more closely mimics what you might need to do in the future, where you won't have a professor guiding your exploration and learning. **

> **Find (at least) three different sources that discuss *differential privacy*.  List your three sources below. Using these three sources, explain in your own words in about one paragraph what differential privacy is, what its benefits are, and what some downsides are. **


Differential privacy is a method to keep individual data private while sharing useful insights from a dataset. It adds random noise to the data or the results of queries, so that the presence or absence of any single person's data doesn't noticeably change the outcome. This makes it hard for anyone to figure out personal details from the shared data. The main benefits are strong privacy protection and the ability to share data insights safely. However, it can be tricky to implement correctly, and the added noise might reduce the accuracy of the data. Balancing privacy and data usefulness is a key challenge.

Sources:
 - [Wikipedia - Differential Privacy](https://en.wikipedia.org/wiki/Differential_privacy)
 - [Towards Data Science - A Gentle Introduction to Differential Privacy](https://towardsdatascience.com/a-gentle-introduction-to-differential-privacy-330434437cdf)
 - [Harvard Privacy Tools - What is Differential Privacy?](https://privacytools.seas.harvard.edu/differential-privacy)



&nbsp;

## Question 2 (2 points)

> **Explain the difference between read_csv and dbConnect().  When would you want to use read_csv(), and when would you want to use dbConnect()? ** 


`read_csv()` is a function used to read data from a CSV file into R as a data frame. It's better for smaller datasets that can be easily stored and manipulated in memory.

`dbConnect()` is used to establish a connection to a SQL database, allowing you to interact with large datasets stored externally. This is useful for working with big data that cannot be efficiently handled in memory.



&nbsp;

## Question 3 (8 points)

> a. **Set up a connection to the database in the Chinook_Sqlite.sqlite file; the rest of this assignment will use this connection. This data is publicly available at https://github.com/lerocha/chinook-database; this link also has come information about the tables included and the relationship between these tables. **

```{r}
chinook_db <- dbConnect(SQLite(), dbname = "Chinook_Sqlite.sqlite")
```



> b. **Use code to output a list of all the tables in the Chinook database. ** 

```{r}
dbListTables(chinook_db)
```


> c. **Set up a code chunk that can run SQL queries for this database. Use code to output all columns of the Album table. **

```{sql, connection = chinook_db}
SELECT * FROM Album
```


> d. **Consider the table Track.  How many rows does it have and how many columns does it have? **

```{sql, connection = chinook_db}
SELECT COUNT(*) FROM Track
```




&nbsp;

## Question 4 (8 points)

> a. **Output just the AlbumId and Name columns from the Track data Table. The names of these two columns in your output should be "Album ID" (with a space between Album and ID) and "Name of Track", respectively.** 

```{sql, connection = chinook_db}
SELECT AlbumId AS 'Album ID', Name AS 'Name of Track'
FROM Track
```



> b. **How many different track names appear in the Track data table? Explain how you know.**

```{sql, connection = chinook_db}
SELECT COUNT(DISTINCT Name) AS num_distinct_names
FROM Track
```



> c. **Arrange the Track table by UnitPrice. When there is a tie in UnitPrice, next arrange by Milliseconds. Be sure you display the UnitPrice, Milliseconds, and Name columns as the first three columns in your answer. Including more columns beyond these three is optional. ** 

```{sql, connection = chinook_db}
SELECT UnitPrice, Milliseconds, Name
FROM Track
ORDER BY UnitPrice, Milliseconds
```



> d. **Arrange the Track data table by Milliseconds, from highest to lowest. There is no need to include any tiebreakers. Be sure you display the Milliseconds and Name columns as the first two columns in your answer. Including more columns beyond these two is optional.** 

```{sql, connection = chinook_db}
SELECT Milliseconds, Name
FROM Track
ORDER BY Milliseconds DESC
```



&nbsp;

## Question 5 (3 points) 

> **Make two new columns in the Track data set, one that displays the length of the track in Seconds and one that displays the length of the track in minutes; these new columns should be calls "Length in Seconds" and "Length in Minutes", respectively.  Note there are 1000.0 milliseconds in one second, and 60.0 seconds in one minute; be sure to use 1000.0 and 60.0 in any calculations you do rather than 1000 and 60. Be sure you display your two new columns, the Milliseconds column, and the Name column as the first four columns in your answer. Including more columns beyond these four is optional.**


```{sql, connection = chinook_db}
SELECT Milliseconds, Name, (Milliseconds / 1000.0) AS 'Length in Seconds', (Milliseconds / 60000.0) AS 'Length in Minutes'
FROM Track
```

&nbsp;

## Question 6 (12 points)

> **In all parts below, make sure your answer displays the column(s) you are using for your filtering and the Name column; including additional columns is optional.**

> a. **Filter the Track data table to find all observations where the GenreID is not 2.**

```{sql, connection = chinook_db}
SELECT *
FROM Track
WHERE GenreId != 2
```


> b. **Filter the Track data table to find all observations where the track Name contains the substring "Love". It is fine if the string you're looking for appears capitalized or lowercase, and it's fine if it appears as a substring of a longer word. **

```{sql, connection = chinook_db}
SELECT *
FROM Track
WHERE Name LIKE '%Love%'
```


> c. **Filter the Track data table to find all observations where the track Name contains "Love" or the track name contains "Hate". It is fine if the strings you're looking for appear capitalized or lowercase, and it's fine if they appear as a substring of a longer word. **

```{sql, connection = chinook_db}
SELECT *
FROM Track
WHERE Name LIKE '%Love%' OR Name LIKE '%Hate%'
```



> d. **Filter the Track data table to find all observations where the track Name contains "Love", "Live", or any other string that has an L, followed by any single character, followed by "ve". It is fine if the strings you're looking for appear capitalized or lowercase, and it's fine if they appear as a substring of a longer word.**

```{sql, connection = chinook_db}
SELECT *
FROM Track
WHERE Name LIKE '%L_ve%'
```



> e. **Filter the Track data table to find all observations where there is no information on the composer of the track.  **

```{sql, connection = chinook_db}
SELECT *
FROM Track
WHERE Composer IS NULL
```



> f. **Filter the Track data table to display rows 1,001 to 1,500.  Your answer should not use the TrackId column; your command should be one that would work on any table, even those without ID columns. **

```{sql, connection = chinook_db}
SELECT *
FROM Track
LIMIT 500 OFFSET 1000
```



&nbsp;

## Question 7 (8 points) 


> a. **Calculate the total Bytes of all tracks in the Tracks data set. **

```{sql, connection = chinook_db}
SELECT SUM(Bytes) AS total_bytes
FROM Track
```



> b. **Calculate the minimum, mean, and maximum milliseconds of the tracks listed in the Track data table.**

```{sql, connection = chinook_db}
SELECT MIN(Milliseconds) AS min_milliseconds, AVG(Milliseconds) AS mean_milliseconds, MAX(Milliseconds) AS max_milliseconds
FROM Track
```



> c. **Do the same calculations as part (b), but only for tracks where composer information is available. **

```{sql, connection = chinook_db}
SELECT MIN(Milliseconds) AS min_milliseconds, AVG(Milliseconds) AS mean_milliseconds, MAX(Milliseconds) AS max_milliseconds
FROM Track
WHERE Composer IS NOT NULL
```



> d. **Do the same calculations as part (b), but separately for each GenreID.**

```{sql, connection = chinook_db}
SELECT GenreId, MIN(Milliseconds) AS min_milliseconds, AVG(Milliseconds) AS mean_milliseconds, MAX(Milliseconds) AS max_milliseconds
FROM Track
GROUP BY GenreId
```



&nbsp;

## Question 8 (16 points) 

> a. (2 points) **What column(s) do the Track and Album tables have in common?  What column(s) to the Album and Artist tables have in common? What column(s) to the Artist and Track table have in common? You should only include columns where the data in the two columns is the same, that is, when these are columns you might want to join by.**

```{r}
# Check the column names of the Track table
dbGetQuery(chinook_db, "PRAGMA table_info(Track)")
```

```{r}
dbGetQuery(chinook_db, "PRAGMA table_info(Album)")
```

```{r}
dbGetQuery(chinook_db, "PRAGMA table_info(Artist)")
```

Answer:
- Track and Album: AlbumId
- Album and Artist: ArtistId
- Artist and Track: ArtistId


> b. (3 points) **Join the Track and Album tables on their common column(s) to create a new table with columns for TrackId, Name (of track), AlbumId, Title (of album), and ArtistId.  Do this join _without_ using the join keyword. **

```{sql, connection = chinook_db}
  SELECT Track.TrackId, Track.Name, Track.AlbumId, Album.Title, Album.ArtistId
  FROM Track, Album
  WHERE Track.AlbumId = Album.AlbumId
```


> c. (3 points) **Produce the same table as the previous part, but this time using the JOIN keyword and the USING keyword. **

```{sql, connection = chinook_db}
  SELECT Track.TrackId, Track.Name, Track.AlbumId, Album.Title, Album.ArtistId
  FROM Track JOIN Album USING(AlbumId)
```



> d. (3 points) **Produce the same tibble as the previous two parts, but this time using the JOIN keyword and the ON keyword. **

```{sql, connection = chinook_db}
  SELECT Track.TrackId, Track.Name, Track.AlbumId, Album.Title, Album.ArtistId
  FROM Track JOIN Album ON Track.AlbumId = Album.AlbumId
```



> e. (2 points) **Explain when you can use the USING keyword, and when you can use the ON keyword. **

- USING: When the column names are the same in both tables.
- ON: When the column names are different in both tables.


> f. (3 points) **Filter the tibble produced in parts (b)-(d) to only include Tracks where the Track does not mention "Walk" and the album title does not mention "Rock".  **

```{sql, connection = chinook_db}
  SELECT Track.TrackId, Track.Name, Track.AlbumId, Album.Title, Album.ArtistId
  FROM Track JOIN Album ON Track.AlbumId = Album.AlbumId
  WHERE Track.Name NOT LIKE '%Walk%' AND Album.Title NOT LIKE '%Rock%'
```




&nbsp;

## Question 9 (13 points)

> a. **(2 points) Make a table displaying all the different ArtistId's that appear in the Album table.If you've done this right, you should get a table with one column and 204 rows. **

```{sql, connection = chinook_db}
  SELECT DISTINCT ArtistId
  FROM Album
```


> b. **(2 points) Make a table displaying all the different ArtistId's that appear in the Artist table. If you've done this right, you should get a table with one column and 275 rows. **

```{sql, connection = chinook_db}
  SELECT DISTINCT ArtistId
  FROM Artist
```



> c. **(3 points) Using your answer to the previous two parts, write a code chunk that determines which ArtistIds appear in both the Album table and in the Artist table. **

```{sql, connection = chinook_db}
  SELECT Album.ArtistId
  FROM Album JOIN Artist ON Album.ArtistId = Artist.ArtistId
```



> d. **(3 points) Using your answers to parts (a) and (b), write a code chunk to determine which ArtistIds appear in the Artist table but not the Album table. ** 

```{sql, connection = chinook_db}
  SELECT Artist.ArtistId
  FROM Artist LEFT JOIN Album ON Artist.ArtistId = Album.ArtistId
  WHERE Album.ArtistId IS NULL
```



> e.  **(3 points) Use an appropriate  join to add the Album information from the Album table to the Artist table, keeping the ArtistId, Artist Name, AlbumId, and Album Title column column. All artists that appear in the Artist table should appear in your output, whether or not they have an album in the Album table.  **

```{sql, connection = chinook_db}
  SELECT Artist.ArtistId, Artist.Name, Album.AlbumId, Album.Title
  FROM Artist LEFT JOIN Album ON Artist.ArtistId = Album.ArtistId
```




&nbsp;

## Question 10 (4 points)

> a. **Explain, in your own words, the difference between WHERE and FILTER(WHERE ), including where in an SQL query they would appear and when you'd want to use one vs. the other.** 

WHERE is used to filter rows before any aggregation occurs, affecting all aggregation functions. FILTER(WHERE ...) is used within an aggregation function to apply a filter only to that specific aggregation. WHERE appears before GROUP BY, while FILTER(WHERE ...) is used within the SELECT clause.


> b. **Explain, in your own words, the difference between WHERE and HAVING, including where in an SQL query they appear and when you'd want to use one vs. the other. **

WHERE is used to filter rows before aggregation, while HAVING is used to filter groups after aggregation. WHERE appears before GROUP BY, and HAVING appears after GROUP BY.
 


&nbsp;

## Question 11 (18 points)

> a. **Create a table that has a column that contains each AlbumId that appears in Track, a column for how many times that AlbumId appears in Track, and a column for the average milliseconds of all tracks on that album. Give your columns appropriate names.** 


```{sql, connection = chinook_db}
  SELECT AlbumId, COUNT(*) AS num_tracks, AVG(Milliseconds) AS avg_milliseconds
  FROM Track
  GROUP BY AlbumId
```


> b. **Modify your answer to (a) so that the count and average only include the observations where the GenreId is 1 (Rock).  That is, you should have a column for AlbumId, a column for the number of tracks on that album that have GenreId 1, and a column for the average length (in milliseconds) of the tracks on that album that have GenreId 1. Your answer should not have any rows for albums with no Rock tracks. Give your columns appropriate names.**


```{sql, connection = chinook_db}
  SELECT AlbumId, COUNT(*) AS RockTrackCount, AVG(Milliseconds) AS AvgMilliseconds
  FROM Track
  WHERE GenreId = 1
  GROUP BY AlbumId
```


> c. **Modify your answer to (b) so that you do have rows for the albums that have no tracks with genre 1; For example, there should be a row in your table for AlbumId 8, with 0 in the Number of Rock Tracks column and NA for the Average Length (in milliseconds) of the rock tracks on that album. **


```{sql, connection = chinook_db}
  SELECT Album.AlbumId, 
         COUNT(CASE WHEN Track.GenreId = 1 THEN Track.TrackId ELSE NULL END) AS RockTrackCount, 
         AVG(CASE WHEN Track.GenreId = 1 THEN Track.Milliseconds ELSE NULL END) AS AvgMilliseconds
  FROM Album
  LEFT JOIN Track ON Album.AlbumId = Track.AlbumId
  GROUP BY Album.AlbumId
```


> d. **Make a table that has three columns: Album Id, Total number of tracks on that album, and total number of rock tracks on that album. **  

```{sql, connection = chinook_db}
  SELECT AlbumId, 
         COUNT(*) AS num_tracks, 
         SUM(CASE WHEN GenreId = 1 THEN 1 ELSE 0 END) AS num_rock_tracks
  FROM Track
  GROUP BY AlbumId
```


> e. **Modify your code from (d) to only keep those AlbumIds that have at least 10 tracks total.**

```{sql, connection = chinook_db}
  SELECT AlbumId, 
         COUNT(*) AS num_tracks, 
         SUM(CASE WHEN GenreId = 1 THEN 1 ELSE 0 END) AS num_rock_tracks
  FROM Track
  GROUP BY AlbumId
  HAVING COUNT(*) >= 10
```


> f. **Modify your code from (d) to also include a column for the Title of each album. **

```{sql, connection = chinook_db}
  SELECT Album.AlbumId, 
         Album.Title, 
         COUNT(Track.TrackId) AS num_tracks, 
         SUM(CASE WHEN Track.GenreId = 1 THEN 1 ELSE 0 END) AS num_rock_tracks
  FROM Album
  LEFT JOIN Track ON Album.AlbumId = Track.AlbumId
  GROUP BY Album.AlbumId
```



&nbsp;

## Question 12 (4 points) 

>  **Using any method you'd like, make a table with two columns: One for Genre (the full Genre name, not the GenreId), and one for how frequently the GenreId for that Genre appears in Tracks. **

```{sql, connection = chinook_db}
  SELECT Genre.Name AS Genre, 
         COUNT(Track.GenreId) AS Frequency
  FROM Track
  JOIN Genre ON Track.GenreId = Genre.GenreId
  GROUP BY Genre.Name
```



&nbsp;

## Question 13 (1 points)

> **Close your connection to the database used throughout this assignment. **


```{r}
dbDisconnect(chinook_db)
```

