---
title: "10-30 Class Notes - SQL, Aggregations, Joins"
author: "Sarah Cannon"
date: "10/30/2024"
output: html_document
---

## Announcements

Homework 8: If using extension time, Final Deadline tomorrow, 11:59pm
Based on class 10/21, 10/23 (Regular Expressions) 

Homework 9: Available tomorrow, due Tuesday, 11/5, 11:59pm 
(can use up to 48 hours extension time) 

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

MAth Tea: Thursday 10/31, 3:30-4:30pm
Chat with faculty and other students about spring courses and more! 
Athenaeum Courtyard

Today is the last day of material for Test 2

Monday, 11/4: Modelling (not on test; on first homework after test)

Wednesday, 11/6: Review for test

Monday, 11/11: Test 2 (Homeworks 6-9)

Questions about upcoming schedule?



## Final Project Rubric

Available on Canvas

Is there anything here that's not clear? 

Overall Guiding Principle: "Does the student demonstrate that they have learned how to work with data this semester?"


## SQL

Questions from preclass video? 

Set up connection to database - for class, sales.sqlite (on Canvas)

```{r}
library(DBI) # to connect to database management systems
library(RSQLite) # To use SQLite in RMarkdown

db_sales <- dbConnect(
          SQLite(), ## Dialect of SQL we will be using
          dbname = "sales.sqlite" ## Name of database and where to find it
)

db_sales
```

```{r}
dbListTables(db_sales)
```

## Joins

Method 1: Use WHERE

Code from Preclass Video:

```{sql, connection = db_sales}
SELECT sales_agent, manager 
  FROM sales_teams
```

```{sql, connection = db_sales}
SELECT Sales_Agent, Account, Deal_Stage
  FROM sales_pipeline
```


```{sql, connection = db_sales}
SELECT pipeline.sales_agent AS "Agent", 
       teams.manager, 
       pipeline.account, 
       pipeline.Deal_Stage
  FROM sales_teams AS teams, sales_pipeline AS pipeline
  WHERE (pipeline.sales_agent = teams.sales_agent) AND 
    pipeline.deal_stage = "Won"
```

Is an inner join


Method 2: Use JOIN keyword

More code, but will conceptually separate the joining and the filtering. 

Now, joining is part of FROM

When joining column(s) have the same name in both tables: Can use USING

```{sql, connection = db_sales}
SELECT pipeline.sales_agent AS Agent, 
       teams.manager, 
       pipeline.account, 
       pipeline.Deal_Stage 
   FROM sales_teams AS teams 
    JOIN sales_pipeline AS pipeline USING(sales_agent)
```

Must have parentheses after USING. 

If the columns we're joining by have different names: Use ON instead of USING

```{sql, connection = db_sales}
SELECT teams.sales_agent AS Agent, 
       teams.manager, 
       pipeline.account, 
       pipeline.deal_stage
   FROM sales_teams AS teams 
      JOIN sales_pipeline AS pipeline 
      ON pipeline.sales_agent = teams.sales_agent
      WHERE Agent = "Carol Thompson"
```

These are both still inner joins, only keep sales agents that appear in both tables. 


### Outer Joins

Right joins, left joins, and full joins are all examples of outer joins.

Use: 
LEFT OUTER JOIN
RIGHT OUTER JOIN (might not always be supported by SQLite)
FULL OUTER JOIN (might not always be supported by SQLite)

Same concept as tidyverse, different syntax!

Example of LEFT OUTER JOIN: 

```{sql, connection = db_sales}
SELECT sales_teams.sales_agent AS Agent, 
  sales_pipeline.deal_stage
  FROM sales_teams LEFT OUTER JOIN sales_pipeline USING (sales_agent) 
```

Is agent Carol Thompson included? 

```{sql, connection = db_sales}
SELECT sales_teams.sales_agent AS Agent, 
    sales_pipeline.deal_stage
 FROM sales_teams 
    LEFT OUTER JOIN sales_pipeline USING (sales_agent)
WHERE Agent = "Carol Thompson"
```

Carol Thompson appears in sales_teams but not in sales_pipeline: she has no deals in the pipeline.  But, she's included here because we did a left join rather than an inner join

This is because it's a left join

To do a right join: 
- If using a robust dialect of SQL: RIGHT OUTER JOIN
- In SQLite: May need to switch the order of your tables

```{sql, connection = db_sales}
SELECT sales_teams.sales_agent AS agent, 
    sales_pipeline.deal_stage
 FROM sales_pipeline 
  LEFT OUTER JOIN sales_teams USING (sales_agent)
 WHERE agent = "Carol Thompson"
```

Doing a right join is the same as doing a left join with the order of your tables swapped. 

To do a Full join: 
- If using a robust dialect of SQL: FULL OUTER JOIN
- In SQLite: May need to do the union of a left join and a right join. 



## Aggregations

Can accomplish the same things as a summarize()

SUM, COUNT, AVG, MIN, MAX. Goes within SELECT, because it's a new column you're making.

```{sql, connection = db_sales}
SELECT SUM(Revenue)
  FROM Accounts
```

To count the number of rows: COUNT(*)

```{sql, connection = db_sales}
SELECT COUNT(*) AS "Number of Deals" 
FROM sales_pipeline
```

Can also be combined with WHERE: 

```{sql, connection = db_sales}
SELECT COUNT(*) AS "Number of Deals Won" 
FROM sales_pipeline
WHERE deal_stage = "Won"
```

Find mean closing value of each deal:

```{sql, connection = db_sales}
SELECT COUNT(*) AS Num_deals_won,
  AVG(close_value) AS Average_Value
  FROM sales_pipeline
  WHERE deal_stage = "Won"
```

### Group By

Same as group_by() from tidyverse. Now summary statistics are computed for each group.

```{sql, connection = db_sales}
SELECT  sales_agent,
    COUNT(*) AS Num_deals_won, 
    AVG(close_value) AS Average_value
  FROM sales_pipeline
  WHERE deal_stage = "Won"
  GROUP BY sales_agent
  ORDER BY Average_value DESC
```

GROUP BY must be after WHERE, before ORDER BY

Works with joins: To group by manager: 

```{sql, connection = db_sales}
SELECT sales_teams.manager, 
    COUNT(*) AS "Number Deals Won",
    AVG(sales_pipeline.close_value) AS "Average Closing Value"
  FROM sales_teams JOIN sales_pipeline USING (sales_agent) 
  WHERE sales_pipeline.deal_stage = "Won"
  GROUP BY sales_teams.manager
```


### FILTER keyword

**DIFFERENT THAN FILTER IN TIDYVERSE!!!!!!!**

For filtering observations *as part of an aggregation*, use FILTER(WHERE(...)) after it (within the SELECT)

e.g., to only count deals with a closing value over 1000

```{sql, connection = db_sales}
SELECT sales_agent,
    COUNT(*) AS "Total Deals Won",
    COUNT(*) FILTER(WHERE( sales_pipeline.close_value > 1000)) AS "Number of deals won over $1000"
  FROM sales_pipeline
  WHERE deal_stage = "Won"
  GROUP BY sales_agent
```

This filtering *only* applies to whatever aggregation function appears immediately before it. 

If you want to do filtering before all aggregation functions: 

To filter rows _before_ aggregations occur: Use WHERE (for all aggregations) or FILTER(WHERE()) for just one aggregation

To filter rows _after_ an aggregation occurs: Use HAVING (use formulas, not a new name you've made) 

```{sql, connection = db_sales}
SELECT sales_agent,
  COUNT(*) AS 'number won'
  FROM sales_pipeline
WHERE deal_stage = "Won"
GROUP BY sales_agent
HAVING COUNT(*) > 200
```

Can so this with the column we made with FILTER(WHERE ) as well. 

```{sql, connection = db_sales}
SELECT sales_agent,
  COUNT(*) AS 'number won', 
  COUNT(*) FILTER(WHERE( sales_pipeline.close_value > 1000)) AS "Number of deals won over $1000"
FROM sales_pipeline
WHERE deal_stage = "Won"
GROUP BY sales_agent
HAVING COUNT(*) FILTER(WHERE( sales_pipeline.close_value > 1000)) > 200
```


### Set Operations

UNION, INTERSECT, MINUS/EXCEPT

To find all accounts, including international

International accounts:

```{sql, connection = db_sales}
SELECT account, office_location AS location
  FROM intl_accounts
```

Domestic Accounts:  (set location manually to USA) 

```{sql, connection = db_sales}
SELECT account, 
    "USA" AS Location
  FROM accounts
```

Combine them with a union: 

```{sql, connection = db_sales}
SELECT account, office_location AS location
  FROM intl_accounts
  
UNION

SELECT account, 
    "USA" AS Location
  FROM accounts
```

Like in R: the two table must have the same columns for this to work! 

To do a full join: do a left join, do a right join (by swapping the order of the tables in your join), and then:  take their union

INTERSECT works extremely similarly to UNION (again, tables must have the same columns), only keeps rows that appear in both tables. 


To find agents in the sales_teams table but not the sales_pipeline table: Use EXCEPT

```{sql, connection = db_sales}
SELECT sales_agent FROM sales_teams
```

```{sql, connection = db_sales}
SELECT sales_agent FROM sales_pipeline
```


```{sql, connection = db_sales}
SELECT sales_agent FROM sales_teams

EXCEPT  

SELECT sales_agent FROM sales_pipeline
```

Is like a set difference (or an anti_join) 

Sometimes, the keyword is MINUS instead (in other dialects of SQL)



## Disconnecting

Good practice: When done, disconnect from the database 

```{r}
dbDisconnect(db_sales)
```

## Order of Keywords

Keywords must appear in the following order (not all of these have to be present)
1. SELECT
2. FROM 
3. WHERE 
4. GROUP BY 
5. HAVING 
6. ORDER BY

## Summary

Inner joins (three options): 
- Use WHERE
- Use FROM ... JOIN ... USING (...) [if joining columns have same name]
- Use FROM ... JOIN ... ON ...

Outer Joins: 
- Use FROM ... LEFT OUTER JOIN ... USING (...) [if joining columns have same name]
- Use FROM ... LEFT OUTER JOIN ... ON ...

Aggregations: 
- Use SUM, MIN, MAX, AVG, COUNT, etc. in SELECT
- Can also use GROUP BY

Different types of filtering:
- Before Aggregation: WHERE ...
- Within one aggregation, but not for all aggregations: FILTER(WHERE ...) within SELECT
- After aggregation: HAVING ...

Set operations: 
- UNION, INTERSECT, EXCEPT/MINUS


## Activity

On Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 

