---
title: "10-9 Class Notes"
author: "Sarah Cannon"
date: "10/9/2024"
output: html_document
---
```{r, results = FALSE, message = FALSE}
library(tidyverse)
```

## Announcements

Homework 6: Available tomorrow, due next THURSDAY (10/17), 11:59pm 
(delayed two days because of fall break) 
Based on class today and 9/30 

No class Monday (Fall Break)

Upcoming Office Hours: 
(no regular office hours during fall break)
- Tuesday 10/15 4-5pm
- Wednesday 10/16 1:15pm-2:15pm 
- Thursday 10/17 2-3pm

Reminder: Thursday 10/17: Last day to drop course without record

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from Preclass Video?


## Sets - From Preclass Video

A _set_ is a collection of elements (order doesn't matter)

Appear in lots of different ways in data science! 

Example 1: The values that appear in a column

Example 2: The rows of a tibble

The _union_ of two sets is all elements that are in at least one of the two sets: A union B = all elements in A or in B

(When set = values that appear in a column: union ~ full join)

The _intersection_ of two sets is all elements that are in both sets: A intersect B = all elements in A and in B

(When set = values that appear in a column: intersection ~ inner join)



## Cartesian Product

For sets A and B, the _Cartesian Product_ A x B is all possible pairs of an element from A and an element from B

Example: A = {1,2}, B = {'a', 'b', 'c'}
A x B = { (1, 'a'), (1, 'b'), (1,'c'), (2, 'a'), (2, 'b'), (2, 'c') }

Number of elements in A x B = (number of elements in A) * (number of elements in set B)

Data Science Example 1: complete(col1, col2)

If A is all possible values in col1 and B is all possible values in col2, then A x B is is all possible pairs of values: the complete function looks at all possible pairs of values, and if some pair isn't in the tibble, it adds that pair to the tibble. 

```{r}
t1_5 <- table1 %>% slice(1:5)
t1_5
complete(t1_5, country, year)
```

This is A x B where A = {Afghanistan, Brazil, China} , B =  {1999, 2000}

Also works for three or more sets: A x B x C = all triples involving an element from A, and element from B, and an element from C.


Data Science Example 2: Joins with duplicate values

```{r}
t5 <- tibble(x = c('p', 'p'), y = c(1, 2))
t6 <- tibble(x = c('p', 'p', 'p'), z = c(4, 5, 6))
t5
t6
```

```{r}
inner_join(t5, t6, by = "x")
```

A = rows in t5 where x = p
B = rows in t6 where x = p

A x B = all combinations of a row from A and a row from B; these give the rows in the join. 



## Filtering Joins

```{r}
library(nycflights13)
```

Airports has so many airports in it  - 1458

```{r}
airports
```

Flights has a lot fewer airports - only those reachable from NYC via a direct flight: 105 destinations

```{r}
flights %>% group_by(dest)
```

What if we wanted to trim down the airports data set and only keep the relevant airports that appear in flights? 

Don't want to actually add any information/data to airports: just want to filter airports tibble based on whether or not destinations appear in flights. 

Filtering joins: Don't add data from one table to another, instead just filter one table based on the values in another table. 


### Semi-joins

semi_join(tibble1, tibble2) keeps all observations (rows) in tibble1 that have a match in tibble2: filters tibble1 according to whether or not values appear in tibble2. 

```{r}
airports
flights
semi_join(airports, flights, by = c("faa" = "dest"))

```

Filtering the airport table, by looking at the faa column, and keeping entries that also appear in the dest column of flights. 

Now we only have 101 airports in data set, the ones that are also destinations in flights. 

Syntax is similar to joins from last class, but really is just filtering! 

Is like an intersection: Takes the intersection of the values in the faa column in airports and the destination column in flights, only keeps those rows of airports that are in this intersection. 

(UNLIKE inner_join because don't add any information from flights). 

Note: 105 destinations in flights but 101 destinations here - will revisit! 

Other examples of semi_joins: 

Suppose you want to only keep flights to the top 10 destinations. 

```{r}
top_dest  <- flights %>% count(dest, sort = TRUE) %>% head(10)
top_dest
```

count(): makes a new table that counts how many items each value appears
sort = TRUE: puts them in order form largest to smallest

How can we filter flights to only show flights to these destinations? 

filter(dest %in% c("ORD", "ATL", "LAX",...))

Easier way: use a semi_join!

```{r}
flights %>% semi_join(top_dest, by = "dest")
```

Works for multiple columns too:  what if you want to keep flights on 30 most popular days? 

```{r}
pop_days <- flights %>% count(month, day, sort = TRUE) %>% head(30)
pop_days
flights %>% semi_join(pop_days, b = c("month", "day"))
```

Common use of semi-join: Make a tibble of values/observations you want to keep (like top_dest or pop_days). Then, use a semi_join of your data set (flights) and this new tibble (top_dest or pop_days) to only keep the relevant observations. 


### Anti-joins

105 destinations in flights

101 destinations in semi_join(airports, flights, by = c("faa" = "dest"))

Must be four destinations that appear in flights but don't appear in airports

How can we find them? 

Let's filter flights to find destinations that don't appear in airports:

anti_join(tibble1, tibble2): keeps all observations in tibble 1 that *don't* have a match in tibble2. Or, removes all observations in tibble1 that *do* have a match in tibble2

```{r}
anti_join(flights, airports, by = c("dest" = "faa")) %>% count(dest)
```

BQN, PSE, SJU: In Puerto Rico
STT: US Virgin Islands

Seems like for flights, domestic destinations include US territories, for airports, it doesn't include US territories.

Anti-joins are useful for diagnosing mismatches. 

Are there any tailnums that appear in flights but don't appear in planes? use anti-join! 

```{r}
planes
flights %>% anti_join(planes, by = "tailnum")
```



## Another set concept: Set difference

The _set difference_ of two sets, A - B or A \ B,  is everythign that is in A but not B

Ex. A = {1,2}, B = {2,3},
A - B = {1}
B - A = {3}

https://www.basic-mathematics.com/images/Difference-of-set.gif

In Data Science: An anti_join is a set difference

A = values in dest column in flights

B = values in faa column of airports

A - B = all values that are destinations in flihgts but don't appear in airports = rows kept by the anti_join(flights, airports)


## R's built-in set functions

R has built-in intersect(), union(), and setdiff() functions: operate on rows

Be very careful to not confuse these built-in functions, which always assume the set you're dealing with is rows, with the more general concepts, which appear in many different settings. 

**ONLY WORK FOR TIBBLES WITH EXACT SAME COLUMNS**

union(t1, t2): all rows that are in at least one of the tibbles
intersect(t1, t2): all rows that are in both of the tibbles
setdiff(t1, t2): all rows that are in t1 but not t2
setdiff(t2, t1): all rows that in t2 but not t1

```{r}
t1 <- tribble(
  ~pet, ~species, 
  "Gracie", "Cat", 
  "Lassie", "Dog"
)

t2 <- tribble(
  ~pet, ~species, 
  "Spot", "Lizard", 
  "Gracie", "Cat", 
  "Tweety", "Bird"
)
t1
t2
```

```{r}
union(t1, t2)
intersect(t1, t2)
setdiff(t1, t2)
setdiff(t2, t1)
```


Possible use for union:

```{r}
Jan_Obs <- tibble(Day = c(2,7,17,19,30), Bird = c("Sparrow", "Sparrow", "Pigeon", "Plover", "Hummingbird"))
Feb_Obs <- tibble(Day = c(3,4,7,12, 27, 28), Bird = c("Hummingbird", "Crow", "Raven", "Hummingbird", "Hummingbird", "Pigeon"))
Jan_Obs
Feb_Obs
```

How to combine into one list? Attempt 1: 

```{r}
union(Jan_Obs, Feb_Obs)
```

Can't tell which birds are from which month! 

Add month information to the tables first: 

```{r}
JanObs2 <- Jan_Obs %>% mutate(month = "Jan")
JanObs2
FebObs2 <- Feb_Obs %>% mutate(month = "Feb")
FebObs2
union(JanObs2, FebObs2)
```


## Summary

A _set_ is a collection of elements (order doesn't matter)

- The _union_ of two sets is all elements that are in at least one of the two sets: 
      A union B = all elements in A OR in B
- The _intersection_ of two sets is all elements that are in both sets: 
      A intersect B = all elements in A AND in B
- The _Cartesian Product_ A x B is all possible pairs of an element of A with an element of B
- The _set difference_ of two sets, A - B or A \ B, is everything that is in A but is not in B

Filtering joins: filter one tibble according to whether or not values appear in a second tibble
- Idea same as filtering, syntax same as mutating joins from last class
- semi_join(t1, t2, by = "name"):  Keep all rows of t1 where the value in the name column also appears in the name column of t2
- anti_join(t1, t2, by = "name"):  Keep all rows of t1 where the value in the name column does not appear in the name column of t2

R has built-in intersect(), union(), and setdiff() functions: operate on _rows_. 
ONLY work on tibbles that have the exact same columns

## Activity

Available on Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 


