---
title: "11-18 Class Notes Linear Models Two Variables"
author: "Sarah Cannon"
date: "2024-11-18"
output: html_document
---


# Announcements

Due Tuesday, 11/19: Homework 10 (modelling, based on classes 11/4 and 11/13) **AND Final Project Proposal** (on Gradescope) 

Homework 10, Question 3: All parts should reference the same model as given in Part (a): 
$new_daily_cases = -100 + 50 * day_number$

Office Hours this week: 

Tuesday **2-2:30pm and 3-3:30pm** (no usual morning office hours)
Wednesday 1:15-2:15pm

Wednesday's preclass video: Review of Logarithms/exponential (not code)

Questions about upcoming schedule?


# Linear Models with Mulitple Variables

```{r}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
```


# Modelling with Two Predictor Variables

Preclass Video: California Childcare costs data set

mc_infant = -310.450 + 8.847 * flfpr_20to64 - 1.224 * pr_f

This model is a plane in three dimensions

For different values of pr_f, the relationship between mc_infant and flfpr_20to64 is a line
This line always has the same slope, no matter the value of pr_f

Questions from preclass video? 


In class: sim4 data set 

```{r}
sim4
```

Goal 1: equation like y = c1 + c2 * x1 + c3 * x2

```{r}
sim4_mod <- lm(y ~ x1 + x2, data = sim4)
sim4_mod
```

Model: y = 0.03546 + 1.92167 * x1 - 2.7852 * x2

Can visualize in Wolfram Alpha: "Plot y = 0.03546 + 1.82167 * x1 - 2.78252 * x2"

Instead, let's understand this model a different way: for different values of x2, what does the relationship between x1 and y look like? Like taking "slices" 

Could pick particular values for x1 and x2 to look at (like we did in preclass video: looked at lines for pr_f = 10 and pr_f = 20)

Or, just look at many values! 

If your data set doesn't have too many unique values (like sim4 - 10 values for x1, 10 values for x2), use data_grid()

data_grid(dataset, col1, col2, ...) returns a data set with all unique combinations of a value from col1 and a value from col2 (and a value from col3, etc.) 

```{r}
data_grid(sim4, x1, x2)
```

Gives us all possible combinations of x1 and x2 in data set. Removes duplicates. 
Let's add on our predictions:

```{r}
sim4_pred <- data_grid(sim4, x1, x2) %>% add_predictions(sim4_mod)
sim4_pred
```

To plot model when x2 = 1: 

```{r}
sim4_pred %>% filter(x2 == 1) %>% 
  ggplot() + 
  geom_line(aes(x = x1, y = pred))
```

geom_line is not drawing a line!  It is connecting the different predicted points, which happen to form a line because we've done a linear model. 

What if we want to draw this line on top of our original data for x2 = 1? 

Need to specify data argument of our geom_point. 

```{r}
sim4_pred %>% filter(x2 == 1) %>% 
  ggplot() + 
  geom_line(aes(x = x1, y = pred)) + 
  geom_point(data = sim4 %>% filter(x2 == 1), aes(x = x1, y = y))
```

We've done this for one slice of our data : the slice x2 = 1.  

We could do this individually for all other slices. 

But, instead, we'll do them all at once! Don't need to draw each line separately.

To plot lines all at the same time: use color! Need to also *group*: within geom_line, this makes a separate line for each value. (Need to do this for numerical variables). 

```{r}
sim4_pred %>% ggplot() + 
  geom_point(data = sim4, aes(x = x1, y = y, color = x2)) +
  geom_line(data = sim4_pred, aes(x = x1, y = pred, color = x2, group = x2))
```

Like with ccc in pre-class video, lines all have same slope, just different intercepts.  

Larger x2 = *smaller* intercept, this makes sense because coefficient of x2 is negative. 

What if slope should change as x2 changes?


## Modeling: Two numerical variables with interactions

For when the slope of the relationship between two variables should change based on the value of a third variable. 

Previous pattern: y = c1 + c2 * x1 + c3 * x2

Now: Add term for x1 * x2

New pattern: y = c1 + c2 * x1 + c3 * x2 + c4 * x1 * x2

How to find this linear model: 

```{r}
sim4_interaction <- lm(y ~ x1 * x2, data = sim4) 
sim4_interaction
```

Equation: y = 0.03546 + 1.82167 * x1 - 2.78252 * x2 + 0.95228 * x1 * x2

For any fixed value of x1, have a linear relationship between x2 and y

For any fixed value of x2, have a linear relationship between x1 and y

But, model is no longer a plane! (because of the x1 * x2 term). Can view in Wolfram Alpha.

https://www.wolframalpha.com/input?i=z+%3D+0.03546+%2B+1.82167+*+x+-+2.78252+*+y+%2B+0.95228+*+x+*+y

Can view models between x1 and y for different values of x2

```{r}
data_grid(sim4, x1, x2) %>% 
  add_predictions(sim4_interaction) %>% 
  ggplot() + geom_line(aes(x = x1, y= pred, color = x2, group = x2))
```

All these lines have different slopes! 

Equation: y = 0.03546 + 1.82167 * x1 - 2.78252 * x2 + 0.95228 * x1 * x2

When x2 = 1: 

y = 0.03546 + 1.82167 * x1 - 2.78252 * 1 + 0.95228 * x1 * 1
 = -2.74706 + 2.77228 * x1

When x2 = -1: 

y = 0.03546 + 1.82167 * x1 - 2.78252 * -1 + 0.95228 * x1 * -1
= 2.81798 + 0.86939 * x1

Relationship between x1 and y changes for different values of x2

Can also see that relationship between x2 and y changes for different values of x1: 

```{r}
data_grid(sim4, x1, x2) %>% 
  add_predictions(sim4_interaction) %>% 
  ggplot() + geom_line(aes(x = x2, y= pred, color = x1, group = x1))
```

Can also add in data points and/or use facets: 

```{r}
data_grid(sim4, x1, x2) %>%
  add_predictions(sim4_interaction) %>%
  ggplot() + 
  geom_line(aes(x = x2 , y = pred, color = x1 , group = x1)) + 
  geom_point(data = sim4, aes(x = x2, y=y, color = x1))
```


Another way to visualize the model: A new geom: geom_tile: 

```{r}
data_grid(sim4, x1, x2) %>% 
  add_predictions(sim4_interaction) %>% 
  ggplot() + 
  geom_tile(aes(x = x1, y = x2, fill = pred))
```

Model predicts: highest values for large x1, small x2; lowest values for small x1, large x2

(Must have data grid and equally-spaced data points for this to work well) 


### seq_range

This data set had nice, evenly-spaced data points.  

This made both our geom_tile and our geom_line with groups work nicely. 

In sim4, x1 took on 10 different values: got 10 lines. 

If x1 takes on 10000 different values, would get 10000 lines - not great! 

To make a data grid, spanning the range of your data, with a fixed number of values: 

```{r}
dg5 <- sim4 %>% data_grid(
  x1 = seq_range(x1, 5), 
  x2 = seq_range(x2, 5)
)
dg5
```

seq_range(x1, 5): returns a list with 5 values, evens spaced between the smallest value for x1 and the largest value for x1.  

Using this data grid: Get 5 lines, or 25 tiles

```{r}
dg5 %>% 
  add_predictions(sim4_interaction) %>% 
  ggplot() + 
  geom_line(aes(x = x1, y = pred, color = x2, group = x2))
```

```{r}
dg5 %>% 
  add_predictions(sim4_interaction) %>% 
  ggplot() + 
  geom_tile(aes(x = x1, y = x2, fill = pred))
```


## Models with one categorical and one numerical predictor

Example: sim3 data set

```{r}
sim3
```

We will try to predict y based on x1 (numerical) and x2 (categorical)

```{r}
count(sim3, x1)
count(sim3, x2)
```

Let's visualize this: 

```{r}
ggplot(sim3) + geom_point(aes(x = x1, y = y, color = x2))
```

Let's try two different models:

1. without iterations between x1 and x2 (for all four colors, model line will have same slope)

2. with interactions between x1 and x2 (for all four colors, model line can have different slopes)

```{r}
mod_wo_int <- lm(y ~ x1 + x2, data = sim3)
mod_int <- lm( y ~ x1 * x2, data  = sim3)
mod_wo_int
mod_int
```

Let's make our predictions for mod_wo_int and mod_int. Could make a separate data grid for each, but will be more convenient to do at same time!

If want to do add_predictions for multiple models at the same time, use gather_predictions:

```{r}
sim3_preds <- sim3 %>% data_grid(x1, x2) %>% gather_predictions(mod_wo_int, mod_int) 

sim3_preds
```

Predictions for each model in different rows: but is actually tidy! model is a variable and so should be in its own column, not as column names (will make plotting easier).

Now, let's draw the model!  

Can do model without interactions first: 

```{r}
sim3_preds %>% filter(model == "mod_wo_int") %>% 
  ggplot() + geom_line(aes(x = x1, y = pred, color = x2)) + 
  geom_point(data = sim3, aes(x = x1, y =y, color = x2))
```

In model without interactions, all lines have same slope (slightly down)

Let's plot both our models at the same time - use facets! 

```{r}
sim3_preds %>%
  ggplot() + geom_line(aes(x = x1, y = pred, color = x2)) + 
  geom_point(data = sim3, aes(x = x1, y =y, color = x2)) + 
  facet_wrap( ~ model)
```

This is an example of why tidy data is good/useful! 

In model with interactions, lines for different values of x2 have different slopes! 

Which is better? 

- Model without interactions is simpler, so if it does a good job, use it! 

- If you have reason to believe the value of the categorical variable (x2) shouldn't change the slope of the interaction between the other two variables, use model without interactions

- Otherwise, model with interactions is often more accurate/a better option

Can check residuals to see.  To calculate residuals for each model, need original data values too - let's add our residuals for the models to the original data set, rather than the data grid.

For multiple models: use gather_residuals, rather than add_residuals

```{r}
sim3 %>% 
  gather_residuals(mod_int, mod_wo_int) %>% 
  ggplot() + geom_point(aes(x = x1, y = resid, color = x2)) + 
  facet_grid(x2 ~ model)
```

facet_grid:  Can break a plot out into facets based on two categorical variables, rather than just one. 

Easier to see patterns in residuals when look at residuals separately for each color.

See here, model with interactions is better: less of a pattern within residuals. 


### Equations of these models

```{r}
mod_wo_int
```

Have new variables x2b, x2c, x2d, which are 0 or 1 according to the value x2 takes on (if all 0, x2 = a)

y = 1.8717 - 0.1967 * x1 + 2.8878 * x2b + 4.8057* x2c + 2.3596 * x2d

Changing the value of x2: only changes the intercept of the relationship between x1 and y

```{r}
mod_int
```

Now, also have products of x1 with all these variables

y = 1.30124 - 0.09302 * x1 + 7.06938 * x2b + 4.43090 * x2c + 0.83455 * x2d 
        - 0.76029 * x1 * x2b + 0.06815 * x1 * x2c + 0.27728 * x1 * x2d
        
When x2 = a: 
y = 1.30124 - 0.09302 * x1 + 7.06938 * 0 + 4.43090 * 0 + 0.83455 * 0 
        - 0.76029 * x1 * 0 + 0.06815 * x1 * 0 + 0.27728 * x1 * 0
  = 1.30124 - 0.09302 * x1

When x2 = d: 

y = 1.30124 - 0.09302 * x1 + 7.06938 * 0 + 4.43090 * 0 + 0.83455 * 1 
        - 0.76029 * x1 * 0 + 0.06815 * x1 * 0 + 0.27728 * x1 * 1
  = 1.30124 - 0.09302 * x1 + 0.83455 + 0.27728 * x1
  = 2.1359 + 0.18426 * x1

Setting x2 = d (x2d = 1, x2c = 0, x2b = 0): adds a correction to *both* the intercept and the slope. 


## Models with two categorical predictors

Extremely similar to models with one categorical predictor

Make one prediction for each category

Before: category = one value of the categorical variable

Now: category = a combination of a value for the first categorical variable and the second categorical variable

Model has terms like: x1_val1 * x2_val2 
Coefficient is contribution to predicted value when x1 takes on val1 and x2 takes on val2

Everything else we learned for categorical models still applies. 



# Summary: Linear modelling with two predictor variables

Model without interactions: lm(y ~ x1 + x2, data = ...)
- If x1 and x2 both numeric: Model is a plane in three dimensions
- if x1 is numeric and x2 is categorical: Model is a line for each category
- The relationship between y and x1 will always have the same slope, regardless of value of x2

Model with interactions: lm(y ~ x1 * x2, data = ...)
- If x1 and x2 both numeric: Model is a curved surface in three dimensions 
        (every slice parallel to an axis is a line)
- if x1 is numeric and x2 is categorical: Model is a line for each category
- The relationship between y and x1 will have different slops for different values of x2
  
New functions: 
- gather_predictions, gather_residuals: used to look at multiple models at the same time
- geom_line with group aes: make a different line for each value of a numeric variable
- geom_tile: Each grid square colored according to value of 3rd variable

wolframalpha.com: can quickly plot equations/surfaces to see what they look like (and more!)


# Activity

On Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 


