---
title: '8-28 Class Notes: R, RStudio Introduction'
author: "Prof. Sarah Cannon"
date: "8/28/2024"
output:
  html_document: default
  pdf_document: default
---

# Announcements

- This template is available on Canvas (Files > Class Files)
- Tip: Save in a folder dedicated to this class (not downloads folder)
- First part of class: R/RStudio Introduction
- Second part of class: Discussion of Data Ethics Case Studies
- Tomorrow: Check and make sure you have access to Wednesday's preclass video on VidGrid
- Due Friday, 12pm (noon): Complete Intro Survey on Gradescope
- Before class on Wednesday (9/4): Watch preclass video on VidGrid
- Due Wednesday (9/4) at 11:59pm on Gradescope: Homework 1 (released tomorrow morning)
- Upcoming office hours (individual meetings or other questions):
  - Tuesday 1:45-3:30pm
  - Wednesday 1:15-2:15pm

Questions? 


# R and RStudio

From last class: 

- Main programming environment we will be using this semester
- Can run commands in console, or in code chunks in an RMarkdown file
- Stored variables show up in environment tab
 
With console: No way to edit previous commands; hard to keep a record of what you've done. 

We won't work in the console much! 

## RMarkdown: 

- A way of organizing code into a single file that you can run in chunks or run all at once (Similarish to Jupyter Notebooks)
- Can insert code chunks in RMarkdown, and run commands there just like you would in the console by clicking green arrow 
- Can run multiple commands at once

```{r}
x <- 3
y <- 2
x+y
x-y
```

- We will mostly use RMarkdown: Can download this template file, and follow along


### Why use RMarkdown? 

Can "publish": Produce a nice document. Called Knitting. For now, we'll knit to html. 
How do you open html files? ANy web broswer, such a chrome, safari, etc. 

Helps with the communication step of the data science process 
Can explain what your code is doing
Also keeps track of/remembers all commands, so you know exactly what you did to your data.


### RMarkdown Commands

NOT "What you see is what you get"

Sections: #, ##, ### (have space after)

Lists: Use - (may need a blank line before your lists)

**bold text**

*italic text*

Math mode: Use dollar signs: $a^2 + b^2 = c^2$

$\frac{1}{3}$

Math on its own line, centered: 

\[y_1 = mx_1 + b \]

LaTeX: a typesetting language for math

Important: no spaces after first dollar sign, before the last dollar sign: 
$a^2$

Most LaTeX math commands can be used in RMarkdown. 


### Knitting to PDF

Produces nicer document; in the long-term (beyond this class), you will probably want to do this

But, need to have LaTeX installed on your computer for this to work (which is why I don't require it)

Steps on Canvas


### Homework: PDF submissions

Gradescope can only accept PDF files for Homework submissions

Option 1: Knit to PDF, submit that PDF

Option 2: Knit to html, Click "Open in Browser". then print the page (as you normally would), but save to PDF instead.


Be sure to assign questions to pages 

https://www.gradescope.com/get_started#student-submission

Don't put your name on your hoomework



### RMarkdown Header

Shouldn't need to change standard header (may look slightly different on different systems)

But, can change title, author, date if you want

Other options reflect how you have recently knit it

Are other options you can set, for example to make document interactive



## R Objects

We've seen variables/objects that are numbers: 

```{r}
x <- 3
```

Just like in python, variables/objects can be more that just numbers 

### Lists

Can combine numbers into a single list: 

```{r}
y <- c(-2, 3, 4)
```

```{r}
y
```
Now y is a list of three numbers: the function 'c' combined them. 
 
Lists don't work quite the same way as python, though. Operations on lists usually operate on each element separately: 

```{r}
y^2
```

It's costly/expensive to change the length of y. It's fairly easy to change one value, though:

```{r}
y[1]
```
(indexing starts at 1, not 0)
```{r}
y[2] <- 6
```

```{r}
y
```

Let's look at: 
```{r}
sqrt(y)
```

The square root of -2 is imaginary : R can't compute it.  NaN denotes "Not a Number", similar to NULL

One of two types of 'empty' values we'll see.


### Tables

R has some built-in variables/data sets. 

```{r}
cars
```

Note: Running a code chunk only shows first ten rows

CAREFUL: Depending on type of table, may show entire table in knitted file!

DO NOT submit homework with entire data sets displayed!

To only show first ten rows, use head function

```{r}
head(cars, 10)
```

To access a single column of a data table, do tablename$columnname - produces a list

```{r}
cars$speed
```

Use head command on long lists too: 

```{r}
head(cars$speed, 10)
```

To access a row: 

```{r}
cars[4,]
```

To access a specific entry: 

```{r}
cars[4,2]
cars$dist[4]
```

To get more information about a built-in function or variable, use ?

```{r}
?cars
```


## R Functions

R has some built-in functions we'll use frequently

c is one example of a function - takes in several values and combines them into a single variable; 

```{r}
y <- c(-2, 5, 3)
y
```

Some other useful functions *that take one argument, a list* are: 

```{r}
mean(y) # sum of the elements divided by the number of elements
sd(y) # Standard Deviations of the elements: how spread out they are
sum(y) # add up the elements
median(y) # when all elements are sorted in numerical order, the middle element
min(y) # smallest element
max(y) # largest element 
```

Comment within a code chunk: Use #

Comments in an RMarkdown document outside a code chunk: <!-- anything you type here won't show up in the knitted file -->

Run current line of code chunk only: CTRL+Enter

Run current code chunk (all of it): Click green arrow, or CTRL+Shift+Enter

These functions need to take a list as the first argument. This gives the wrong answer: 
```{r}
mean(-2, 3, 5)
```

Instead, combine value into a list: 

```{r}
mean(c(-2, 3, 5))
```

What happens when one value is NaN? 

```{r}
z <- sqrt(y)
z
mean(z)
```
If one value is NaN, so is the whole mean/sum/etc. 


### min vs. pmin

min - the smallest value across all arguments/lists
pmin - parallel min - does *pairwise* comparisons between two (or more) lists

```{r}
l1 <- c(3,6,10)
l2 <- c(1, 7, 8)

#Finds the minimum value across both l1 and l2
min(l1, l2)

# Compares l1 and l2 entrywise
pmin(l1,l2)

```

Note: for pmin, l1 and l2 should be the same length

```{r}
l3 <- c(4,5)
pmin(l1, l3)
```

(because l1 is longer than l3, entries of l3 are repeated in order until it's the same length as l1)
(Don't need to know exactly what happens when l1 and l3 aren't the same length, but should know that you shouldn't do this!)

Should only use pmin on lists of the same length

Note: same for max, pmax, but for maximum instead of minimum. 


## Defining our own functions, function arguments

Can also define your own functions: 
(We mostly won't do this, but it's useful here to understand how arguments of functions work in R)

```{r}
add2 <- function(a = 2, b = 4) {
  s <- a + 2*b
  return(s)
}
```

Note: 2 and 4 are the default values for a and b 
(what gets used if these arguments aren't specified)

Note function appears in Environment panel (it's an R object!)

```{r}
add2(3,7)
```

When one of the arguments isn't specified when the function is called, the default values are used

```{r}
add2(3)
```

Also assumed: The parameter 3 passed to the function was the value for a. 

Arguments are assumed: to be given in the order from the function definitions

Can manually specify if you want

```{r}
add2(b = 3)
add2(b = 3, a = 5)
```

Best practice: Specify which argument is which (a = 5, b = 7, etc.) 



## R Libraries

R has some built-in functions, commands, variables

To add more, use libraries! 

If you try to use a function in a library without installing/loading that library first, get an error
```{r}
#ggplot(data = cars) + geom_point(mapping = aes(x = speed, y = dist))
```

The very first time you use a library on a machine, need to install it (need to use quotes around library name)
Can do this in a code chunk, or in the console
Only need to run this command once

```{r}
# install.packages("tidyverse")
```

Note: install.packages() commands will cause errors when knitting.
If do in code chunk, comment them out after running once. 

After installing it, every time you open R, need to load this library. No quotation marks. 

```{r, message = FALSE}
library(tidyverse)
```

Loading tidyverse can output a lot of message in the knitted file; change start of code chunk to {r, message = FALSE} to hide this. 

Now, can use any functions from the tidyverse package: 

```{r}
ggplot(data = cars) + geom_point(mapping = aes(x = speed, y = dist))
```

In RMarkdown file, libary() command must come *before* any code chunks using thigns from that library 


Packages might have functions, data, or both. 

```{r}
# install.packages("nycflights13")
library(nycflights13)
flights
```


## Troubleshooting/Debugging

What to do if your code doesn't work/doesn't produce the answer you expected/your file won't knit? 

This is a skill that you'll get better at with practice! 

- Read through code to look for any obvious typos

- Compare your code to similar code chunks we've written in class; are there any differences? 

- Look at any error messages you're receiving; what are they trying to tell you? 
  - Knitting errors will show up in the bottom left, with line numbers

- Start simple and build up: Start with a simple command that you know works, and gradually build up until you have the whole code chunk.  

- Try to figure out exactly which part of the command is causing the issue


## Checking your answers

When finishing a question, always pause and ask, "Is this answer reasonable"? 

Ex. You're looking at data of all flights departing a single airport in a year, and you find that the average departure delay is 389 minutes.  Probably not right! 

Go back and look at the question, make sure you're answering all parts of it.


## Data Ethics Activity

In small groups.


## Data Ethics Key Takeaways







