---
title: "Class Notes 9-11 - Transforming Data"
author: "Sarah Cannon"
date: "9/11/2024"
output: html_document
---


## Announcements

Homework 2: If using extension hours, final deadline Thursday 11:59pm

Homework 3: Posted tomorrow morning, due Tuesday 11:59pm (all about data transformations)

Upcoming Office Hours: 
- Monday 1:15-2:15
- Tuesday 10:30-11:30 and 2-3
- Wednesday 1:15-2:15

Homework Collaboration: If want to be connected with classmates, send me an email; will put students in touch with each other

Template for class notes: On Canvas, Files > Class Materials > 9-11 Class Notes Template

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?


# Transforming Data

```{r, message = FALSE}
library(nycflights13)
library(tidyverse)
```

## Preclass Video: Pipes and Filter Helper Functions

```{r}
filter(flights, between(dep_delay, 0, 10))

flights %>% filter(between(dep_delay, 0, 10))
```

Questions from preclass video?


## Grouping and Summarizing

What we're *not* doing today: using the summary() function

```{r}
summary(cars)
```

We're using the *summarize* function (or summarise)

At first, nothing happens!

```{r}
summarize(flights)
```

We didn't say what to include in the summarize.

Here's an example: 

```{r}
summarize(cars, avg_distance = mean(dist))
```

Makes a new table, single column called avg_distance, calculated by doing mean(dist)

Format: 

summarize([data table], [name for new column 1] = [how new column 1 is calculated])

Note: Format for new columns similar to format for mutate

OR, use pipes! 

```{r}
cars %>% summarize(avg_distance = mean(dist))
```

Can add multiple columns at the same time: 

```{r}
cars %>% summarize(avg_distance = mean(dist), 
                   avg_speed = mean(speed), 
                   min_speed = min(speed)
                   )
```

within summarize, use min, not pmin
within mutate, use pmin, not min
min = min of entire column(s)
pmin = entry-wise min across columns

Let's summarize flights: 

```{r}
flights %>% summarize(avg_dep_delay = mean(dep_delay),
                      med_dep_delay = median(dep_delay))
```
We have NA values. 

But this makes sense - we already know there's flights with dep_delay = NA, and we also know that taking a mean of a bunch of numbers, where some are NA, will  output NA.

```{r}
mean(flights$dep_delay)
```

Another option: remove the NA values when taking the mean

BE CAREFUL WHEN DOING THIS: Check why values are NA. Simply removing NA values without understanding why they're there can lead to bias in data set. 

For flights: removing NA values means we will find the average departure delay *of the flights that took off* excluding the flights that didn't take off.

Add na.rm = TRUE inside mean/median/max/sum/etc. to remove any NA values from the calculation. 

```{r}
flights %>% summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE), 
                      med_dep_delay = median(dep_delay, na.rm = TRUE))
```

 **BE CAREFUL** removing NA values from calculations. 

```{r}
?mean
```

Recall from Monday: Can use sum(is.na([some column])) to find number of NA values. This can also be done within summarize.  

```{r}
flights %>% summarize(num_no_takeoff = sum(is.na(dep_delay)))
```

Summing a list of TRUE/FALSE values give you the number of TRUEs

But, if there are NA values, get NA

Can also remove NA values from sums to get number of TRUE values even when there are NA values: 

```{r}
flights %>% summarize(num_flights_delayed_more_than_60 = 
                                sum(dep_delay > 60, na.rm = TRUE))
```


## group_by

Takes a tibble and breaks it down into groups. 

At first, when using this command, nothing appears to change in our table! 

```{r}
group_by(flights, month)
by_month <- flights %>% group_by(month)
by_month
```

by_month looks the same as flights! Group information at top

If your data is all in a box, think of group_by like writing on the outside of the box: "whenever you open this box and try to summarize the data, use these groups." 

Data itself remains unchanged, there is just some extra information attached to it. 

Now, when use summarize(), treats each group separately: 

```{r}
by_month %>% summarize(avg_delay = mean(dep_delay, na.rm = TRUE))
```

Because the by_month data set had been grouped by month, there are 12 groups in the data, summarize calculates for each group separately. 

This is the first command we've learned that uses a part of the data table that we can't see.

For everything we've done so far using summarize(), if you group first, can get values for each group. 

```{r}
flights %>% group_by(month) %>% 
  summarize(num_delayed_more_than_60 = sum(dep_delay > 60, na.rm = TRUE))
```


### n()

Can use other functions besides mean

Can use anything you could use for summarize: median(), min(), max(), sum(), etc. 

A new function: n(): This counts how many rows there are; when data is grouped, counts how many rows there are in each group. 

```{r}
summarize(by_month, 
            avg_delay = mean(dep_delay, na.rm = TRUE), 
            med_delay = median(dep_delay, na.rm = TRUE), 
            num_flights_this_month = n())
```

Note: Generally the n() function should be used within summarize, mutate, filter,... (on it's own it's not that interesting). 

Can use n() in mutate as well: 

```{r}
by_month %>% 
  mutate(number_of_flights_this_month = n())
```

There were 27004 flights schedule in January, so every row for January has this value; rows for other months will have other values. 

(most useful as an intermediate step, for example if you now want to do additional calculations...)


### rank()

We learned rank() on Monday: ranks rows according to a given column

When data is grouped: ranks within groups

```{r}
by_month %>% mutate(order_within_month_by_Dep_delay = rank(dep_delay))
```

The first row is the 17759.5th least delayed flights in January. 

(remember: can get ranks with 0.5 because of ties)


### Avoiding intermediate values using pipes; combining transformations

by_month is an intermediate value - didn't really need it, can use pipes instead! 

```{r}
flights %>% 
  group_by(month) %>% 
  summarize(delay = mean(dep_delay, na.rm = TRUE))
```

This is how you'll most often want to use group_by: with pipes, before summarize

Suppose we are interested in how flight delays are related to the distance of the flights. 

Just making a scatterplot of this isn't very useful  (also slow)

```{r}
ggplot(data = flights) + geom_point(mapping = aes(x = distance, y = arr_delay))
```

Instead, let's find the average delay for each destination

```{r}
flights %>% 
  group_by(dest) %>% 
  summarize(num_flights = n(), 
            dist_to_dest = mean(distance, na.rm = TRUE), 
            dest_delay = mean(arr_delay, na.rm = TRUE))
```

Let's only look at destinations with at least 20 flights: easy to add this in using pipes! 

```{r}
flights %>% 
  group_by(dest) %>% 
  summarize(num_flights = n(), 
            dist_to_dest = mean(distance, na.rm = TRUE), 
            dest_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  filter(num_flights >= 20)
```

Filtering before summarize: reference columns in original data set
Filtering after summarize: reference columns made in summary table (like num_flights)

Let's plot it! Using pipes, we can pass this data table directly into ggplot. 

```{r}
flights %>% 
  group_by(dest) %>% 
  summarize(num_flights = n(), 
            dist_to_dest = mean(distance, na.rm = TRUE), 
            dest_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  filter(num_flights >= 20) %>% 
  ggplot() + geom_point(mapping = aes(x = dist_to_dest, 
                                      y = dest_delay, 
                                      size = num_flights), alpha = 0.3)
```

Observation: as distance gets longer, delay gets shorter 

Key ideas: 
- can use pipes to pass things to ggplots
- for making useful plots, sometimes you want to group and summarize first


### new geom: geom_smooth

This draws a curve that shows the general trend of the data

Often drawn of top of a geom_point, using same mapping 

```{r}
flights %>% 
  group_by(dest) %>% 
  summarize(num_flights = n(), 
            dist_to_dest = mean(distance, na.rm = TRUE), 
            dest_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  filter(num_flights >= 20) %>% 
  ggplot() + geom_point(mapping = aes(x = dist_to_dest, 
                                      y = dest_delay, 
                                      size = num_flights), 
                        alpha = 0.3) + 
      geom_smooth(mapping = aes(x = dist_to_dest, 
                                      y = dest_delay))
```

Shaded region: standard error ("uncertainty" in curve)

To remove: se = FALSE

To draw a line rather than a curve: method = 'lm'

```{r}
flights %>% 
  group_by(dest) %>% 
  summarize(num_flights = n(), 
            dist_to_dest = mean(distance, na.rm = TRUE), 
            dest_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  filter(num_flights >= 20) %>% 
  ggplot() + geom_point(mapping = aes(x = dist_to_dest, 
                                      y = dest_delay, 
                                      size = num_flights), 
                        alpha = 0.3) + 
            geom_smooth(mapping = aes(x = dist_to_dest, 
                                      y = dest_delay), se = FALSE, method = 'lm')
```


### Grouping by multiple columns

Note: can also group by multiple columns: 

```{r}
flights %>% group_by(month, day) %>% 
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE))
```

For each month/day combo in the data set, have a row in your summarized table

All other aspects of summarizing are the same

Example Question: Which route (destination/origin pair) has the longest average delay? 

```{r}
flights %>% group_by(origin, dest) %>% 
  summarize(average_arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  arrange(desc(average_arr_delay))
```

Answer: EWR -> CAE (Newark, NJ to Columbia, SC)


## geom_bar with stat = "identity"

What if you want a plot that shows the average arrival delay by month? 

```{r}
average_monthly_delay <- by_month %>% summarize(avg_delay = mean(dep_delay, na.rm = TRUE))
average_monthly_delay
```

Before: bar charts for *counts* only (or, when using position = "fill", for proportions)

Now: bar charts where lengths/heights are *data values*

Every geom has a default stat: for geom_bar, it's count

```{r}
?geom_bar
```

To use explicit data values rather than counts, need to change the stat argument

Use stat = "identity" to use explicit values in the data rather than counting to get the heights of your bars. 

```{r}
ggplot(data = average_monthly_delay) + 
  geom_bar(mapping = aes(x = month, y = avg_delay), stat = "identity")
```

Need to set an x and a y aesthetic (no longer have y as counting by default) 

Be careful: if an x-value has multiple y-values, will just add them in the plot!  Using stat = "identity" could produce a plot that's not what you want. 

```{r}
# This code makes an example tibble
# You don't need to know how it works
gpas <- tibble(college = c("CMC", "HMC", "Scripps", "Pitzer", "Pomona", 
                           "CMC", "HMC", "Scripps", "Pitzer", "Pomona"), 
               major = c("DS", "DS", "DS", "DS", "DS", 
                         "Math", "Math", "Math", "Math", "Math"), 
               GPA = c(3.5,3.4,3.7,3.9,3.2,3.6,3.5,3.8,3.9,3.4))
gpas
```

```{r}
ggplot(data = gpas) + geom_bar(mapping = aes(x = college, y = GPA, fill = major), stat = "identity")
```

Better: make a summary table first of exact values you want to see in the plot, and then use geom_bar with stat = "identity"

Alternately: using geom_bar with stat = "identity" is the same as using geom_col() 

```{r}
ggplot(data = average_monthly_delay) + 
  geom_col(mapping = aes(x = month, y = avg_delay))
```


## A new geom: geom_line

geom_line: connects consecutive data points with a line. 

Only use when your data is a series/sequential (x-values that have a meaningful order, one y-value per x-value).

Can be used in place of geom_point (if one y-value for each x-value) or in place of geom_bar with stat = "identity" (the the x-values have a meaningful order) 

```{r}
ggplot(data = average_monthly_delay) + 
  geom_line(mapping = aes(x = month, y = avg_delay))
```

```{r}
economics
ggplot(data = economics) + 
  geom_line(aes(x = date, y = psavert))
```

Mostly used for time series, when x-values are dates, hours, years, etc. to show how a y-variable is changing over time. 

Won't work is data is not sequential: 

```{r}
# This is not a good plot
ggplot(data = mpg) + geom_line(mapping = aes(x = cty, y = hwy))
```

Note: geom_line does not draw a line. Draws many line segments, connecting each data point to the next. 

Note: Can also use geom_smooth to draw a curve on top of your line plot. 



## Reminder: Using pipes to select columns

Recall: In Homework: use  select to put relevant columns first

From last class: 
```{r}
flights_newcols <- mutate(flights, 
       gain = dep_delay - arr_delay, 
       speed = distance / air_time * 60)

select(flights_newcols, gain, dep_delay, arr_delay, speed, distance, air_time, everything())
```

This is much easier with pipes!

```{r}
mutate(flights, 
       gain = dep_delay - arr_delay, 
       speed = distance / air_time * 60) %>% 
  select(gain, dep_delay, arr_delay, speed, distance, air_time, everything())
```



# Summary

summarize(): use to compute mean, median, sum, min, etc. of data columns

group_by(): Doesn't change any data values, attaches extra information to data set that indicates certain future calculations should be done for each group separately


geom_smooth(): draws a curve that approximates the trends in the data

geom_bar() with stat = "identity": For making a bar chart where lengths/heights of bars are values given in the data (not counts)

geom_line(): draws line segments connecting consecutive data points; only use for sequential data


# Activity

On Canvas, Files > Class Material > 9-11 Activity

If finish all regular questions, move on to bonus question.

If finish bonus question, can leave: check in with Prof. Cannon before leaving

Feel free to talk to/work with others around you.

