---
title: "Class Notes 9-16 - Advanced ggplots, data transformations"
author: "Sarah Cannon"
date: "9/16/2024"
output: html_document
---

```{r}
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

## Announcements

Homework 3 Due tomorrow night at 11:59pm on Gradescope as a PDF. Assign *all* pages to a question! https://help.gradescope.com/article/fnbn2knmq1-how-can-i-submit-my-homework-as-a-pdf

Upcoming office hours: 

- **Tuesday 3-4:30pm** (no morning office hours) 
- Wednesday 1:15-2:15pm

Template for class notes: On Canvas, Files > Class Materials > 9-16 Class Notes Template. 

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from pre-class video?


## Visualizing with ggplot

```{r}
library(tidyverse)
```

Previously: Many types of plots! 
Today: A more principled viewpoint about when to use which kinds of plots


## Single-variable plots

First, always look at variables individually to understand spread/range/distribution.

Categorical variables:

- bar chart
- composition plot 

Numerical variables:

- boxplot

```{r}
ggplot(data = mpg) + geom_boxplot(aes(x = cty))
```

- histogram 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), color = "black", fill = "white")
```

Can change binwidth: 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), color = "black", fill = "white", binwidth = 5)
```

Choosing the right number of bins takes practice!  Depends on what questions you're asking about your data. 

Here's a bad value for binwidth: 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), color = "black", fill = "white", binwidth = 0.75)
```

Why are there so many gaps between 10 and 20? 

These gaps are artificial! 
All cty values are integers
 
```{r}
mpg
```
 
When the bins have width less than one, some bins can't have any data values in them! 

If data is integers: Always set the bindwith to be an integer. 

If data is rounded to 2 decimal places: Binwidth should not be more than 2 decimal places.  And so on... 

A good binwidth for this plot: 1 or 2 or 5

Too few bins: hides some interesting features. 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), color = "black", fill = "white", binwidth = 8)
```

Alternately: specify number of bins rather than binwidth: 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), color = "black", fill = "white", bins = 10)
```

Usually better to specify binwidth - here don't have control over width, don't know if it's an integer, etc. 

But, in certain situations, might make sense to specify number of bins


## Plots with Two Variables

Two categorical variables: 

- bar charts with fill

Two numeric variables: 

- scatterplot (geom_point)
- line graph (geom_line - when data is sequential) 
- geom_smooth (usually drawn on top of one of the previous two - shows general trends, not actually data)

One numeric and one categorical variable

- geom_bar with stat = "identity" (when there's one numeric y-value for each categorical x-value)
- boxplots
- histograms with facets


### Boxplots with Two Variables

Used with one numeric and one categorical variable

Instead of one boxplot for the whole data set, make a separate boxplot for the data within each category

Instead of just using an x aesthetic, use a y aesthetic too! 

```{r}
ggplot(data = mpg) + geom_boxplot(mapping = aes(x = cty, y = class))
```

Each category has different IQR, different whiskers, different median, different outliers, etc. 

Can also color the boxes differently if you'd like: 

```{r}
ggplot(data = mpg) + geom_boxplot(aes(x = cty, y = class, fill = class))
```

Reminder: fill = color inside a share, color = outline of a shape


### Facets

This separates your plot out into multiple panels, one for each value of a categorical variable.

Most useful when there's only a few values for your categorical variable. 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), binwidth = 2, color = "black", fill = "white") + 
  facet_wrap(~ drv)
```

Format: + facet_wrap( ~ cat_col_name)

Can be used for any type of plot to split up according to a categorical variable, not just histograms

For example, with geom_point: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv)
```

Not as useful when too many facets: 

```{r}
ggplot(data = mpg) + 
  geom_histogram(aes(x = cty), binwidth = 2, color = "black", fill = "white") + 
  facet_wrap( ~ class)

```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap( ~ class)
```


## More than two variables

Add onto existing plots using features such as color, facets, size, etc. 

Example: scatterplot: for two numeric variables, but can add information about additional categorical variables with color, shape, or facets, or additional numerical variables using size or color. 

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = drv))
```

Can also add additional features to geom_smooth, geom_line: 

In a geom_smooth, can get a separate curve for each value of a categorical variable. 

For example, for drv: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), se = FALSE)
```

Or, can remove color from geom_smooth to just have one curve: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy), se = FALSE, color = "black")
```


In geom_line, can get a different line graph (collection of line segments) for different values of a categorical variable.

Example from bonus activity: 

```{r}
# making the built-in data set airquality more amenable to plotting 
# (will learn how/why to do this later!) 
airquality <- airquality %>% mutate(Month = factor(Month))
airquality

ggplot(airquality) + geom_line(mapping = aes(x = Day, y = Wind, color = Month ))
```

To sum up: When plotting, think first about how many variables you want to represent in your plot and whether they are categorical or numeric.  Then, consider what you want to communicate/what's important in your situation: simplicity vs. detail, separate trends vs. comparing trends, etc. 


## Themes, Colors

Can change how your plot looks.

There are many themes you can use in a plot.

For example, make it black-and-white (useful if printing): + theme_bw()

```{r}
ggplot(data = mpg) + geom_point(aes(cty, hwy)) + theme_bw()
```

Remove grid lines, simple style: + theme_classic()

```{r}
ggplot(data = mpg) + geom_point(aes(cty, hwy)) + theme_classic()
```

Are many more themes!  https://ggplot2.tidyverse.org/reference/ggtheme.html


### Colors: Discrete Values (Categorical variables) 

Can also change the default color choices.  Can help with readability, accessibility, etc. 

For example: Most common form of color-blindness is red-green color-blindness. 
R's default when there are three colors includes both red and green.

To change, use: + scale_color_manual( values = list_of_colors) 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), se = FALSE) + 
  scale_color_manual( values = c("blue", "red", "black"))
```

R has built-in colors you can reference: 

```{r}
colors()
```


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), se = FALSE) + 
  scale_color_manual(values=c("violet" , "wheat2" , "bisque"))
```

Or: can specify color with six-digit hexadecimal code
For example, blue is "#0000FF"

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), se = FALSE) + 
  scale_color_manual(values=c("#0000FF", "darkolivegreen", "plum"))
```


This changes colors; to change fill, use scale_fill_manual()

Can also use scale_fill_grey() and scale_color_grey() to make plots greyscale (usually use with theme_bw() or theme_classic, etc.)

Built-in color sets: RColorBrewer package; see  https://r-graph-gallery.com/38-rcolorbrewers-palettes.html

```{r}
#install.packages("RColorBrewer")
library(RColorBrewer)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), se = FALSE) + 
  scale_color_brewer(palette = "RdYlBu")
```


### Color Gradients (for numeric variables) 

For example:  Color gradient by default is light blue to dark blue; another color scale might make more sense. 

```{r}
ggplot(mpg) + geom_point(aes(x = cty, y = hwy, color = displ)) 
```

Can use scale_color_gradient(): specify low and high values

```{r}
ggplot(mpg) + geom_point(aes(x = cty, y = hwy, color = displ)) + 
  scale_color_gradient(low = "blue", high = "red")
```

Can use scale_fill_gradient() for fill instead of color.

Are methods for specifying a gradient with 3 colors, or even more colors: 
scale_color_gradient2(), scale_fill_gradient2() 
scale_color_gradientn(), scale_fill_gradientn()


## Overview

A good visualization should: 

- Pick out the aspects of the data that are important to you. 
- Allow you to see multiple dimensions/variables simultaneously in the same plot
- Allow comparisons across different characteristics of the data set
- Be clearly labelled and easy to understand. 


## Additional Data Transformations

We'll cover some more advanced transformations that build on what we learned last week. 

### Weighted Averages

Sometimes, you want to take a *weighted* average - not all values have same weight. 

```{r}
grades <- tibble(item = c("Test 1", "Test 2", "Quiz", "Homework", "Class Engagement", "Final Project"), score = c(90, 95, 80, 95, 100, 80), weight_perc = c(16, 16, 10, 30, 8, 20))
grades
```

Don't just want to average your grades; want to *weight* them appropriately. 

Use weighted.mean function

```{r}
?weighted.mean
```

```{r}
summarize(grades, final_grade = weighted.mean(score, weight_perc))
```

A common use: when you want to average rates, but the rates come from populations of different sizes. 

```{r}
c <- tibble(county = c("USA", "Canada", "Mexico"), 
            rate = c(0.5, 0.1, 0.2), 
            population_millions = c(337, 40, 131))
c
```

Incorrect: this is not overall rate across the three countries

```{r}
mean(c$rate) # This gives the wrong overall rate
```

Correct: Need to weight according to country's population

```{r}
weighted.mean(c$rate, c$population_millions)
```

Can use weighted.mean in a summarize, or on it's own - it just depends



### Sum/Mean of T/F

We already saw: When sum a TRUE/FALSE vector (excluding NA values, if appropriate), answer is number of TRUEs.

We used it for sum(is.na(column_name)) to count number of NA values. 

Can use it more generally. 

First, make a column with TRUE/FALSE values in it. 

```{r}
library(nycflights13)

flights %>% mutate(is_very_delayed = ifelse( dep_delay > 60, TRUE, FALSE))
```

But, we're just putting TRUE in our new column if the condition is TRUE, and FALSE if the condition is FALSE (and NA if dep_delay is NA). 

Instead, could just directly put this TRUE/FALSE value in new column

```{r}
flights %>% mutate(is_very_delayed = (dep_delay > 60))
```

Now, can sum this new column (removing NA values) to get number of very delayed flights. 

```{r}
flights %>% mutate(is_very_delayed = (dep_delay > 60)) %>% 
  summarize(num_very_delayed = sum(is_very_delayed, na.rm = TRUE))
```

What if want to know what *fraction* of flights that take off are very delayed? 

Will start with an example without any NA values. 

What fraction of cars in mpg have cty > 20? 

```{r}
mpg %>% 
  mutate(cty_greater_than_20 = (cty > 20) ) %>% 
  summarize( num_greater_than_20 = sum(cty_greater_than_20), 
             frac_greater_than_20 = sum(cty_greater_than_20)/n() )
```

Adding elements and dividing by the number of them is taking a mean! 

```{r}
mpg %>% 
  mutate(cty_greater_than_20 = (cty > 20)) %>% 
  summarize(num_greater_than_20 = sum(cty_greater_than_20), 
            frac_greater_than_20 = mean(cty_greater_than_20))

```
Taking the mean of a TRUE/FALSE column gives you the *fraction* of the values that are TRUE. 

Ex. c(T, T, F, T, F, F)

sum(c(1,1,0,1,0,0))/6 = mean( c(1,1,0,1,0,0)) = 0.5

What fraction of flights that took off were very delayed?  Just use mean! 

```{r}
flights %>% mutate(is_very_delayed = (dep_delay > 60)) %>% 
  summarize(num_very_delayed = sum(is_very_delayed, na.rm = TRUE), 
            frac_very_delayed = mean(is_very_delayed, na.rm = TRUE)
            )
```

Or, could skip intermediate step of making new column: 

```{r}
flights %>% 
  summarize(num_very_delayed = sum(dep_delay > 60, na.rm = TRUE), 
            frac_very_delayed = mean(dep_delay > 60, na.rm = TRUE ))
```

To get the fraction of data satisfying some CONDITION, do: summarize( frac = mean(CONDITION))

Here CONDITION is anything that evaluates to TRUE/FALSE (anything you could use in filter())

Example: Fraction of flights that go to Atlanta
CONDITION: dest == "ATL"

```{r}
flights %>% summarize(frac_to_atl = mean( dest == "ATL" , na.rm = TRUE))
```

All of this can be done with group_by() as well: 

```{r}
flights %>% group_by(origin) %>%  
  summarize(frac_to_ATL = mean(dest == "ATL", na.rm = TRUE))
```


## Summary

* = plots discussed today

One-variable plots: 
- bar chart (categorical)
- composition plot (categorical)
* boxplot (numeric)
* histogram (numeric)

Two-variable plots: 
- scatterplot (two numeric)
- line graph (two numeric, sequential data, one y-value for each x-value)
- bar chart with fill (two categorical)
- bar chart with stat = "identity" (one categorical, one numeric - one y-value for each x-value)
* boxplot (one categorical, one numeric)
* histogram with facets (one categorical, one numeric)

To incorporate a third variable: 
- size (numeric)
- shape (categorical, <= 6 values)
* color (numeric or categorical)
* facets (categorical, few values)

Can change how a plot looks using themes or alternate color choices

Weighted averages: weighted.mean()

To count number of rows satisfying CONDITION: sum(CONDITION)
To get fraction of rows satisfying CONDITION: mean(CONDITION)


## Activity

On Canvas, Files > Class Material > 9-16 Activity

If finish all regular questions, move on to bonus question.

If finish bonus question, can leave: check in with Prof. Cannon before leaving

Feel free to talk to/work with others around you.


