---
title: "9-23 Class Notes: Importing, Strings"
author: "Sarah Cannon"
date: "9/23/2024"
output: html_document
---

```{r}
knitr::opts_chunk$set(fig.width=4, fig.height=2)
```

# Announcements

Homework 4 Due tomorrow night at 11:59pm on Gradescope as a PDF. Assign *all* pages to a question! https://help.gradescope.com/article/fnbn2knmq1-how-can-i-submit-my-homework-as-a-pdf

Upcoming office hours: 

- Tuesday 10:30-11:30 and 2-3pm
- Wednesday 1:15-2:15pm

Template for class notes: On Canvas, Files > Class Materials > 9-23 Class Notes Template. 

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from pre-class video?


# Load Libraries

```{r, message = FALSE}
library(tidyverse)
library(nycflights13)
```


# Dates, Times, Date/Times

To turn strings into dates or times: 

1. figure out the pattern of the date or time, using abbreviations

e.g. "23 Apr 2021" is "%d %b %Y"

(Tip: keep table of abbreviations handy)

2. parse the date or time 

```{r}
parse_date("23 Apr 2021", "%d %b %Y")
```

Like last class, first argument is thing to parse: can be a string, a list of strings, or a column of a data set that is of type <chr> 

Questions from preclass video?

## datetime

Often, are together in one column! 

Standard format looks like: "2024-09-23 16:15"

```{r}
parse_datetime("2024-09-23 16:15")
```

If you have date-times in other formats, just need to specify the format. 

See abbreviations at: https://r4ds.hadley.nz/datetimes#tbl-date-formats

```{r}
parse_datetime("23 Sep 2024, 04:15pm", "%d %b %Y, %I:%M%p")
```

Can parse entire columns rather than just a single datetime. 

```{r}
datetime <- tibble(datetime = c("Sep 21 2021, 09:35am","Aug 02 2021, 09:37am"))
datetime %>% mutate( new_datetime = parse_datetime(datetime, "%b %d %Y, %I:%M%p"))
```

<S3: POSIXct>: data type for standard datetimes 

(POSIX: set of standards for maintaining compatibility between operating system)



# Strings

Data type <chr> (not a single character)

A _string_ is an ordered list of symbols from some alphabet

(Have used them, but haven't really learned about them in depth, that's what we're going to do now)

In R, to indicate a string, surround it by either single or double quotes

```{r}
string1 <- "This is a string." 
string2 <- 'This is also a string.'

string1
string2
```

Output is always in double quotes. Good practice: get into habit of using double quotes.


## Encodings, parse_character

How are non-English characters stored? 

Everything in a computer is stored in binary! As 0s and 1s.

For English characters, there's an agreed upon way to turn letters into 0s and 1s (ASCII)

For other languages, there have been a lot of different ways for turning characters into binary.

Today: Mostly use a particular method/encoding called UTF-8 (unicode), a standard encoding (in some other countries, may still use other formats)

*Key Takeaway*: If you're getting gibberish in your data, try using other encodings and the parse_character function. 


```{r}
x <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"
x
#writeLines(x)
```

```{r}
parse_character(x, locale = locale(encoding = "Shift-JIS"))
```

If you don't know encoding, are methods to try to guess (BONUS activity) 



## Escape Characters

For strings, output is not always the same as input. 

If you want to use single quotes in your string: As long as the outside quotes are double quotes, this is fine. 

```{r}
string3 <- "What a 'great' idea!"
string3
```

If want to use double quotes in your string? 

```{r}
# string4 <- "She said, "Hello"" # THis doesn't work
```

One solution: use single quotes on the outside

```{r}
string4 <- 'She said, "Hello"'
string4
```

For inside quotes: denoted with escape character \" : a backslash in a string tells you the next character should be treated differently. 

(Note: To make escape characters show up in your knitted file: Use two backslashes \\ ) 

To actually see string, with the quotes, use writeLines

```{r}
writeLines(string4)
```

An _escape character_ in a string is a single backslash followed by a symbol. Together, they have a different meaning inside a string. 

\": Double quote that is part of your string rather than ending your string
\': Single quote that is part of your string rather than ending your string
\n: New line: The letter n isn't in the string, but instead denotes the start of a new line.
\t: Inserts a tab in your string 
\u: Beginning of a unicode character

```{r}
"\u00e1"
"\u00b0"
"\u263a"
```

Newline character: 

```{r}
string5 <- "This is my string! 
This is the second line of my string!"
string5
writeLines(string5)
```


## Helpful String Functions

R has built-in string functions, we'll use tidyverse alternatives 
(specifically, the stringr package, which is part of the tidyverse)

Most start with str_

We've already learned two: str_detect, str_c

For all of them: first argument is a string, a list of strings, or a column in a tibble of type <chr>

### str_detect

Returns true or false if a pattern is found. 

```{r}
artists <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-17/artists.csv") 
filter(artists, str_detect(artist_nationality, "American"))
```


### str_c

Concatenates/combines strings. 

```{r}
zip_codes <- tibble(Zip = c("91711", "20322", "93782", "78392", "87639", "47628", "20874"), 
                   PlusFour = c("3452", "3009", "8473", "8762", "2563", "5416", "5726"))
zip_codes
zip_codes %>% 
  mutate(full_zip_code = str_c(Zip, "-", PlusFour)) %>% 
  select(full_zip_code)
```

Also works on individual strings:  

```{r}
str_c("a", "b", "c")
str_c("Claremont", " ", "McKenna")
```

If pass a vector/list, it won't concatenate them! 

```{r}
str_c( c("x", "y", "z"))
```

When given lists of things, str_c concatenates them *entrywise*. 

```{r}
str_c( c("a", "b", "c"), c("d", "e", "f"))
```

This is actually what we're seeing above with zipcode: each column is a list, they get concatenated entrywise, and the resulting list is put in the new column. 

When lists are not the same length: Get an error

```{r}
str_c( c("pre", "post"), c("run", "workout", "swim"))
```

Doesn't work with NA: The entire result becomes NA

```{r}
str_c("Claremont", NA, "McKenna")
```

### str_replace_na 

If actually want NA to show up: use str_replace_na(). When applied to a string, does nothing.  When applied to NA, by default, replaces an NA with the string "NA"

```{r}
str_replace_na("Hello")
str_replace_na(NA)
```

Or, can be used to replace NA with something else

```{r}
str_c("Claremont", str_replace_na(NA, replacement = "?"), "McKenna")
```

Can use str_replace_na within mutate/with columns, too. 

```{r}
flights %>% 
  mutate(Announcement = str_c( "This flight takes off at ", 
                               dep_time)) %>% tail()

flights %>% 
  mutate(Announcement = str_c( "This flight takes off at ", 
                               str_replace_na(dep_time, replace = "XXXX"))) %>%
  tail()
```

For flights with a departure time, that departure time is displayed; for flights where the departure time is NA, it displays the time XXXX. 


### str_length

Finds the length of strings

```{r}
str_length("abc")
```

```{r}
str_length("") # Empty string, length 0
```

```{r}
str_length("a\nb") # Escape characters like \n only count as one character
```

First argument is what string(s) to apply the function to: can apply it to a single string, can apply it to a list of strings, can apply it to a string column in a tibble 

Can use it within a filter, for example:

```{r}
filter(diamonds, str_length(cut) >= 5)
```


### Changing Case

str_to_upper, str_to_lower, str_to_title

```{r}
s <- "The cat in the Hat"
str_to_lower(s)
str_to_upper(s)
str_to_title(s)
```

Why? If data is in inconsistent format

```{r}
pets <- tribble(
  ~name, ~species,
  "Sparky", "DOG",
  "Fido", "Dog",
  "Fluffy", "Cat", 
  "Lassie", "DoG",
  "Patches", "cat",
)
pets
pets %>% filter(species == "dog")
```

One solution: change case so it's consistent within the column

```{r}
pets %>% mutate(species2 = str_to_title(species)) %>% filter(species2 == "Dog")
```


### Substrings

str_sub: pulls out a substring from a given starting position to a given ending position (including both the start and the end) 

```{r}
colors <- c("blue", "red", "green")
colors
str_sub(colors, 2, 4)
```

Indexing in R starts at 1. 

This code picks out the substring starting with the second character and ending with the fourth character (or earlier, if the string doesn't have four characters)

Can use negative numbers to count back from the end. This finds the last character of each string:

```{r}
str_sub(colors, -1,-1) #-1 means the last character
```

Arguments to str_sub: 

- thing you want to operate on (string, list of strings, <chr> column in a tibble)
- starting position (inclusive)
- ending position (inclusive)

Used a lot in data cleaning when you have multiple important values in the same column. 

Here's a way to use str_sub to capitalize strings (if you forget str_to_title):

```{r}
colors
str_to_upper(str_sub(colors, 1, 1))
str_sub(colors, 1, 1) <- str_to_upper(str_sub(colors, 1, 1))
colors
```

str_sub only looks at positions, not what's in them. 

For picking out substrings based on what's in the strings, use regular expressions; we'll learn that in a few weeks. 

## Tips from Past Students

- Read through class notes after class/before starting homework
- Make yourself a summary/cheatsheet to reference while doing homework
- Separate the concepts (e.g. only keep certain rows of a data set, turn strings into times) from the commands (e.g. filter, parse_time)
- Start assignment earlier
- Work with classmates


## Summary 

To parse dates, times, or datetimes: 

1. Identify pattern using abbreviations %y, %m, etc. 
2. Use appropriate parse_ function, passing pattern as the second argument

Strings: 

- Can be encoded in different ways
- escape characters: \ changes the meaning of whatever character comes after it
- str_length: find length of a string
- str_c: combine multiple strings into one string
- str_to_upper, str_to_lower, str_to_title: changes case of letters in strings
- str_sub: picks out a substring by giving start and end positions


## Activity

On Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 

