---
title: '9-25 Class Notes: Tidy Data'
author: "Sarah Cannon"
date: "9/25/2024"
output:
  html_document: default
  pdf_document: default
---

```{r}
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

# Announcements

Homework 4: If using extension hours, final deadline Thursday 11:59pm

Homework 5: Posted tomorrow, due Tuesday 11:59pm (based on class this week) 

Today is the last day of material that will be on Test 1 
- All material form the start of the semester thorugh today (including today)
- Everything on Homeworks 1-5

Next Monday 9/30: New Material (not on Test 1, will be on first homework after Test 1) 

Next Tuesday 10/1: Homework 5 due

Next Wednesday 10/2: Review for Test 1 in class (no preclass video)

Monday 10/7: Test 1 

(No Homework due Tuesday 10/8)

Upcoming Office Hours: 
- Monday 1:15-2:15
- Tuesday 10:30-11:30 and 2-3
- Wednesday 1:15-2:15

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from preclass video?


# Tidy Data

```{r, message = FALSE}
library(tidyverse)
```

"Tidy data" is NOT a synonym for clean data. Tidy data is a specific data format. 

In tidy data: 

- Each row corresponds to an observation
- Each column corresponds to a variable
- Each entry only contains one value. 

Nearly all the data sets we've seen so far are tidy! 

```{r}
artists <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-17/artists.csv") 
artists 
# Each row is an occurrence of an artist in an edition of a textbook
```

Why do you want tidy data?

- Advantage to picking a consistent way of storing data - easier to learn tools to work with it
- All packages in tidyverse (ggplot2, dplyr, etc.) are built specifically to work with tidy data
- Advantage of placing variables in columns because of R's vectorized nature.

Tidy Data:

```{r}
table1
```

Untidy Data:

```{r}
table2
```


## Turning entries into column names: pivot_wider

For example, to make table2 tidy.  Want to take the 12 rows and change them back into 6 observations by combining rows with the same country/year. 

Use pivot_wider(): remove the count and type variables, introduce a cases variable and a population variables

pivot_wider has three main parameters: 

- the data set (usually via a pipe) 
- names_from: the column in the current table that holds the entries that we'll turn into column names. ( For table2, the type column) 
- values_from: the existing column that has the values we'll put in the new columns we're making (for table2, the count column)

```{r}
table2 %>% pivot_wider(names_from = type, values_from = count)
```

Now, each variable has its own column, and each row has only values, not variable names. 

If there are more than 2 distinct values in the names_from column, the resulting tibble will be wider. 

Another Example:

```{r}
stud_info <- tribble(
  ~Name, ~College, ~Info, ~Value, 
  "Student A", "CMC", "GPA", 3.8, 
  "Student B", "Scripps", "GPA", 3.7, 
  "Student C", "Pitzer", "GPA", 3.72, 
  "Student D", "CMC", "GPA", 3.66, 
  "Student A", "CMC", "Graduation", 2022, 
  "Student B", "Scripps", "Graduation", 2024, 
  "Student C", "Pitzer", "Graduation", 2023, 
  "Student D", "CMC", "Graduation", 2023, 
  "Student A", "CMC", "Age", 22, 
  "Student B", "Scripps", "Age", 20, 
  "Student C", "Pitzer", "Age", 19, 
  "Student D", "CMC", "Age", 20, 
)
stud_info
```

This data is not tidy: Each row should be one student. We don't want observations about student A split across multiple rows. 

Each column should only correspond to a single variable, here the Value column actually has three different variables in it - GPAs, graduation years, and Ages. 

Rule of Thumb: Does combining all the values in a column (e.g. by a sum or an average) make sense? If not, these values shouldn't be in the same column, and you probably want to pivot wider. 

```{r}
pivot_wider(stud_info, names_from = Info, values_from = Value)
```



## Turning column names into entries: pivot_longer

What about when column names are values for a variable rather than the name of a variable?  

```{r}
table4a
```

This data is not tidy: 1999 and 2000 are not variables, they are really values for another variable, year. Want these values to be within a column, not as column names. 

pivot_longer turns column names into entries, which removes columns and adds rows (makes your tibble longer). Often needed for cateogrical data (or census data) 

pivot_longer needs 4 parameters: 

- data set that you want to pivot_longer
- which of the existing columns are values, not variables
- make up a name for the new variable to put the current column names in (for table4a - year) 
- make up a name for the new variable to put the values in (here, cases)


```{r}
table4a %>% pivot_longer(c(`1999`, `2000`), names_to = "YEAR", values_to = "CASES")
```

Note the quotes around "year" and "cases": these are new names we're creating, not objects we've already made, so need to use strings. 

Also, quotes around 1999 and 2000: anytime variable names start with numbers, have spaces, or have special characters, need to use quotes around them (often, back quotes ` ). 

Should also turn year column into integers. 


Another example: 

```{r, echo = FALSE}
stud_fav_colors <- tribble(
~ College, ~Red,  ~Green, ~Blue, ~Purple, ~Orange, ~Yellow, ~Other, 
"Claremont McKenna College", 36, 75, 67, 24, 57, 98, 99, 
"Harvey Mudd College", 84,72,19,45,92,54,11, 
"Pitzer College", 36,76,21,33,23,56,44, 
"Pomona College", 98,56,45,32,47,56,88, 
"Scripps College", 34,56,28,73,49,87,33,
)
```

```{r}
stud_fav_colors
```

This data is not tidy! Column names are really different values that a variable favorite_color could take on. 

Rule of Thumb: Might you want to combine/average/add values in different columns? Then they should be in the same column. Use pivot_longer. 

```{r}
stud_fav_colors %>% pivot_longer( c(Red:Other),
                                  names_to = "Favorite_Color", 
                                  values_to = "Num_Students")
```

This is now tidy! 

Now easier to make plots showing the relationship between college and favorite color

Census Data:

```{r}
cd <- read_csv("Census_disability_data_2019_B18101.csv", skip = 1)
cd
```

Not tidy! Want to pivot_longer

```{r}
cd2 <- cd %>% pivot_longer( `Estimate!!Total:`:`Margin of Error!!Total:!!Female:!!75 years and over:!!No disability`, names_to = "Description", values_to = "Count" ) %>% 
  select(-id, -`Geographic Area Name`)
```

If column names have special characters, spaces, or start with a number, use quotes or backquotes around them. 

Still not tidy, but closer. 


## Separating Data 

Needed when table has values of two variables in one entry: 

```{r}
table3
```

In table 3, variable "rate" actually has two values: cases, then /, then population. 

Or, census data above has same problem (after pivoting): Multiple descriptors are part of the same string

Can use str_sub if parts always have the same number of characters (zipcodes)

Another option: use separate(). Parameters are: 

- data set
- variable whose values need to be separated (rate)
- A list with the new names of the variables you want to create when you do the separation.  (cases, population) 

```{r}
table3 %>% separate(rate, into = c("CASES", "POPULATION") ) 
```

Note: separate() guessed that / was the separator, and was correct. 

separate() treats the two entries as if they're character strings, rather than values

Can tell it to guess what the data types of the new columns should be: 

```{r}
table3 %>% separate(rate, into = c("cases", "population"), convert = TRUE)
```


For census data: 

```{r}
cd2 %>% separate(Description, into = c("measurement_type", "details") , sep = "!!", extra = "merge" )
```

Need to specify separator: use argument sep, here sep = "!!" 

It cut off everything after the second !!. Two possible solutions: 

- give more column names in into argument 
- use extra = "merge" 

Now can pivot_wider to get estimate and margin of error in separate columns:

```{r}
cd2 %>% separate(Description, 
                 into = c("measurement_type", "details") , 
                 sep = "!!", 
                 extra = "merge" ) %>% 
  pivot_wider(names_from = measurement_type, values_from = Count)
```

Note: Estimates and margins of error are different values, so don't want them in the same column. 

(Would it make sense to sum all the estimates and all the margins of error together? No!)


## Uniting Data

unite()

Opposite of separate

```{r}
table5
```

Similar to str_c within mutate

More on Bonus Activity


## Ambiguous Cases for Mulitple Values 

Does a datetime column have multiple values in it?  

It depends! 


Should first and last name be together in the same column? 

It depends! 


## Summary

"Tidy data" is NOT a synonym for clean data. Tidy data is a specific data format. 

In tidy data: 

- Each row corresponds to an observation
- Each column corresponds to a variable/factor
- Each entry only contains one value. 

pivot_wider: used to turn entries into column names. Has 3 main arguments: 
- The data set (frequently via a pipe)
- names_from: the existing column that holds the entries we'll turn into new variables
- values_from: the existing column that has the values we'll use for these new variables

pivot_longer: used to turn column names into entries. Has 4 main arguments: 
- data set (frequently via a pipe)
- which of the existing column names are values, not variables. 
- the name of the new variable to move the column names to. 
- the name of the new variable to move the column values to.

separate: used when a column has multiple values in a single entry
- data set (frequently via a pipe)
- variable whose values need to be separated (rate)
- new names of the variables (cases, population)


## Activity

On Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 
