---
title: "9-30 Class Notes -  Keys, Joins"
author: "Sarah Cannon"
date: "9/30/2024"
output: html_document
---


## Announcements

Today: New Material (not on Test 1, will be on first homework after Test 1) 

Tomorrow 10/1: Homework 5 due

Wednesday 10/2: Review for Test 1 in class (no preclass video)

Test 1 covers: 

- All material form the start of the semester through last Wednesday 9/25
- Everything on Homeworks 1-5
- Preclass videos, Activities, but not BONUS activities

Monday 10/7: Test 1 

(No Homework due Tuesday 10/8)

**Notes on Homework Grading**: 
-0.01 means there is a comment I want you to see, but it will not affect your overall grade. 
Make sure you *correctly* assign questions to pages. We need to see *all* pages containing your answer, not necessarily the question. 
Make sure your code is not cut off (especially if knitting to PDF).

Upcoming Office Hours: 

- Tuesday 10/1 10:30-11:30 and 2-3
- Wednesday 10/2 1:15-2:15
- Monday 10/7 *10:30-11:30am* and 1:15-2:15pm

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from preclass video? 


## Relational Data

When data is spread across multiple tables

It is the relationships between these tables, not just the tables themselves, that are important.

```{r, message = FALSE}
library(tidyverse)
library(nycflights13)
flights
weather
planes
```

### Assumptions

1. Data is tidy (if not tidy: tidy it first!)

2. Order of rows does not matter.  If it does matter: Make a new column called order

If the order of the rows is important, add it explicitly to the data set. 

```{r}
politicians <- tribble( ~ First_Name, ~Last_Name, ~Party, 
                        "Tammy", "Baldwin", "Democrat", 
                        "John", "Barrasso", "Republican", 
                        "John", "Kerry", "Democrat",
                        "Marsha", "Blackburn", "Republican"
                        )
politicians
```

```{r}
politicians2 <- politicians %>% mutate(order = row_number())
politicians2
```

Now, even if we reorder this tibble, we still know the original order. 

Can put the tibble rows in any order, and still have the necessary information. 

```{r}
politicians2 %>% arrange(First_Name)
```

After this, order doesn't matter and Assumption 2 is met. 


3. There are no identical rows. If there are identical rows: 

```{r}
names <- tibble(name = "Smith, John", age = c(47, 47))
names
```

Do not know if there was a data entry error, or were actually two John Smiths. 

This is where documentation of data collection and other data practices are really helpful! 

When collecting data, helpful to create a unique ID for each observation/row

(A) If there really should be two John Smiths: Create unique IDs to distinguish them.

One way: use row_number()

```{r}
names %>% mutate( id = row_number()) %>% select(id, everything())
```

Here, id column is a key, more specifically, it's a *surrogate key*, a key you made up to distinguish the observations. 

Now, there are no rows that are entirely identical, assumption 3 is met. 

(B) If this data was unintentionally duplicated and there really shouldn't be two John Smiths: Use distinct()

```{r}
names
distinct(names)
```

Now, there aren't duplicate rows. 

Knowing which option you should use requires knowledge about your data! 



## Keys, Foreign Keys

In a data frame, a _key_ is a collection of columns/variables that uniquely identify each observation.  That is, it's a set of columns such that no two rows that have the same values in all of the columns in your key. (order of columns doesn't matter). 

*Key: A precise definition, not an intuition*

Key for flights: { month, day, hour, carrier, flight } (primary key)

Key for planes: {tailnum}

```{r}
planes
```

Check that it is a valid key: 

```{r}
planes %>% count(tailnum) %>% filter(n > 1)
```

tailnum also appears in flights: it connects the two data sets. 

tailnum is not a key (or a primary key) in flights: some tailnums appear multiple times in flights. 

```{r}
flights %>% count(tailnum)
```

A *foreign key* in a data table is a variable or collection of variables that form a key in a *different* table. 

tailnum is a foreign key for flights, because it's a key in a different table, planes. 

A variable can be both a primary and a foreign key: 

carrier is part of the primary key for flights, and it's also a foreign key for flights because it's a key in a different table, namely the airlines table

```{r}
airlines
```

Another example: origin is part of the primary key for weather. (origin, month, day, hour)

```{r}
weather
airports
```

Origins is also a foreign key for the weather table because it's a primary key in a different table, airports (where it has a different name, faa - that's fine!)


## Mutating Joins

Used to bring data from one table into another table.

Called *mutating joins* because the end result is new columns being created, like mutate() does

First, make a smaller data set for us to work with today: 

```{r}
flights2 <- flights %>% select(month:day, hour, origin, dest, tailnum, carrier)
flights2
```

We have carrier using two-letter abbreviation. What if we want full name of carrier? That's in the airlines tibble

```{r}
airlines
```

Combine them using left_join: 

```{r}
left_join(flights2, airlines, by = "carrier" )
```

"left" because we include all data values that appear in the left table. 

This is the most commonly used join, but there are others as well. 

Will learn how joins work using small tibbles. 

```{r}
t1 <- tibble(letter = c('a', 'b', 'c'), color = c('red', 'green', 'blue'))
t2 <- tibble(letter = c('a', 'b', 'd'), sound = c('low', 'medium', 'high'))
t1
t2
```

### Inner Joins

Join type 1: Inner join: Creates observations for all values that appear in both tables. 

```{r}
inner_join(t1, t2, by = "letter")
```

Only keeps values for "letter" that appear in both tables. Other observations (c,d) are thron away.  Have attached both the color and sound tables. 


### Outer Joins - Left, Right, Full

Join type 2: Left Join: Keeps all values that appear in the first (left) tibble, inserts NA when that value doesn't appear in the second (right) tibble. 

```{r}
left_join(t1, t2, by = "letter")
```

Join Type 3: Right Join: Keeps all values from the second (right) table, inserts NA if that value doesn't appear in the first (left) table.  

```{r}
right_join(t1, t2, by = "letter")
```

Join type 4: Full join: Keeps all values from *both* tables, inserting NA as appropriate

```{r}
full_join(t1, t2, by = "letter")
```

- left_join, right_join, and full_join are all *outer joins*

Can (and usually do) use pipes: 

```{r}
t1 %>% left_join(t2, by = "letter")
```

Note that the differences of the different types of joins depend on what different values are in the column(s) you're joining by, not how many times those values are there. 


## Duplicate Values

So far this assumed all the things we were joining by were keys, that is, they uniquely identified a row in _both_ tables.  This may not always be the case. 

If there are duplicate values in one of your two tibbles: attached information is duplicated eachtime the value appears (in all types of joins - inner, left, right, full)

```{r}
t3 <- tribble(
  ~class, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)
t4 <- tribble(
  ~class, ~val_y,
     1, "class1_info",
     2, "class2_info"
)
t3
t4
left_join(t3, t4, by = "class")
```

In this case, class is not a key in t3, it is a key in t4. Because it's a key in t4, it's a foreign key in t3 (because it's a key in a different table)

Flights example: tailnum is a primary key in planes, foreign key in flights.  

```{r}
flights2 %>% left_join(planes, by = "tailnum")
```

This has nothing to do with the fact that it's a left join: would be the same for an inner, right, or full join. 

```{r}
left_join(t3, t4, by = "class")
right_join(t3, t4, by = "class")
inner_join(t3, t4, by = "class")
full_join(t3, t4, by = "class")
```

These all give the same table, because t3 and t4 have the same set of values in the class column (1 and 2).

Note: In practice, the column(s) you're joining by should be a key in at least one of your two tables. 

What happens if there are duplicate values in both tibbles? (This is often an error) 

R doesn't know what data belongs with what; get all possible combinations

```{r}
t5 <- tibble(x = c('a', 'a'), y = c(1, 2))
t6 <- tibble(x = c('a', 'a', 'a'), z = c(4, 5, 6))
t5
t6
left_join(t5, t6, by = "x")
```

(a is a key in neither tibble) 

(because the values for x appears in both tables, left_join, right_join, inner_join, and full_join all give the same table) 

When do left, right, inner, and full joins give different answers? 
- When there are values that appear in the joining column(s) in one table but not the other

When does information get duplicated? 
- Whenever a value in one table appears multiple times



## More joining options

When a key is not a single column: 

```{r}
flights2
weather
flights2 %>% left_join( weather, by = c("day", "month", "hour", "origin") )
```

Links rows that have the same values in all four columns in the two tibbles

If don't specify by parameter, R will use all columns that appear in both tibbles

```{r}
flights2 %>% left_join(weather)
```

Look at message to see what it's joining by; always check that it's joining things in the way you want. 

This is called a *natural join*.

What if a column in one table has a different name in another table?

Ex. origin/dest column in flights "lines up" with faa column in airports. 

```{r}
airports
flights2 %>% left_join(airports, by = c("dest" = "faa"))
```

This does this join by comparing/lining up the dest column in flights2 with the faa column in airports

Format: by = c("name of joining column in left tibble" = "name of joining column in right tibble")

Can Specify multiple column names: 

by = c("col1_left" = "col1_right", "col2_left" = "col2_right"), etc. 


# Merge

The base R function merge can accomplish the same tasks

dplyr/tidyverse     Base R
inner_join(x, y):   merge(x, y)
left_join(x, y):    merge(x, y, all.x = TRUE)
right_join(x, y):   merge(x, y, all.y = TRUE),
full_join(x, y):    merge(x, y, all.x = TRUE, all.y = TRUE)

Better to use tidyverse versions because: 
- tend to be faster in practice
- naming conventions similar to SQL, so this makes SQL easier to learn later


# Summary 

Assumptions: 

1. Data is tidy 
2. Order of rows does not matter  
3. There are no identical rows

Key = column(s) whose values uniquely identify each observation/row

Foreign key = column(s) that are a key in a *different* table

Mutating joins: 
- inner_join: keep all observations where joining value(s) appear in *both* tibbles: in the left tibble *AND* the right tibble
- left_join: keep all observations from *left* tibble
- right_join: keep all observations from *right* tibble
- full_join: keep all observations of joining value(s) from *either* tibble: in the left tibble *OR* the right tibble

Duplicate values: joined data is duplicated each time! 

Columns you're joining by should usually be a key in at least one of the data tables. 



# Activity

On Canvas

If finish regular questions, move on to bonus questions.

If finish bonus questions, check in with Prof. Cannon, then can leave. 

Can work individually or with classmates. 


