---
title: "Class Notes 9-9-2024 Data Transformation with dplyr"
author: "Sarah Cannon"
date: "9/9/2024"
output: html_document
---


# Announcements

Homework 2 Due tomorrow (TUESADAY) night at 11:59pm on Gradescope as a PDF. Assign *all* pages to a question! https://help.gradescope.com/article/fnbn2knmq1-how-can-i-submit-my-homework-as-a-pdf

Upcoming office hours: 

- Tuesday 10:30-11:30am and 2-3pm
- Wednesday 1:15-2:15pm

Homework Collaboration: If want to be connected with classmates, send me an email; will put students in touch with each other

Template for class notes: On Canvas, Files > Class Materials > 9-9 Class Notes Template

QCL one-on-one appointments and drop-in schedule: https://www.cmc.edu/qcl/mentoring

Questions about upcoming schedule?

Questions from preclass video?



# Transforming Data 

```{r, message = FALSE}
library(nycflights13)
library(tidyverse) # Or: dplyr, which has a subset of what's in the tidyverse
```

```{r}
flights
```

## Filter

```{r}
lateflights <- filter(flights, dep_delay > 60)
lateflights
```

What is really happening: making a TRUE/FALSE list for every row

```{r}
head(flights$dep_delay > 60, 150)
```

Keep only rows corresponding to TRUE

Is this all flights that did not depart within 60 minutes of their scheduled time?

NO!

```{r}
tail(flights)
```

We missed the flights that didn't take off! 

Data often has missing values. E.g., a survey respondent didn't fill out age.

Appears in data frame as NA: "Not Available"
(Similar to NaN: two types of unavailable values) 

Data set we'll be using in activity has some NA values: 

```{r}
starwars
```

Cannot compare NA's to values, such comparisons result in NA: 

```{r}
NA > 60
```

```{r}
NA + 70
```

```{r}
tail(flights$dep_delay > 60, 150)
```

filter() only keep values that are TRUE, not values that are FALSE or NA

We can explicitly check to determine whether something is NA

```{r}
is.na(3)
is.na(NA)
```

Can apply to vectors as well: 

```{r}
x <- c(56, 2, NA, 42)
is.na(x)
```

Find all flights that didn't take off: 

```{r}
filter(flights, is.na(dep_time))
```

There were 8255 flights with no departure time listed

Would need to know more about how the data was collected to be sure these were actually cancelled/never took off.

Should also check that dep_delay is NA exactly when dep_time is NA, etc. 

Could there be a flight with dep_time = NA, but dep_delay != NA? 
(Can use ! to negate True/False values)

```{r}
x <- c(3,5,7,NA,9)
is.na(x)
!is.na(x) # checking whether things are not NA
```

(This applies to any TRUE/FALSE vectors, including others that might appear in a filter())

```{r}
filter(flights, is.na(dep_time), !is.na(dep_delay))
filter(flights, !is.na(dep_time), is.na(dep_delay))
```

(Recall: in filter, comma acts like AND)

Are no rows, so there are no flights where dep_time is NA but dep_delay is not NA, and vice versa

Can check arr_time, arr_delay, air_time too. 

All flights that did not take off within an hour of their scheduled departure time: 

```{r}
not_on_time <- filter(flights, dep_delay > 60 | is.na(dep_delay))
not_on_time
```

(More examples of is.na() in textbook)


## Using arrange to order rows

Will sort rows from lowest to highest: 

```{r}
arrange(flights, year, month, day)
```

- first argument = data set
- second argument = main column to sort by 
- third argument = column to sort by if there are ties in the first argument
- fourth argument = column to sort by if there are ties in the first and second argument
etc. 
(only need to have one column to sort by, rest are optional)

If you want to sort from high to low, surround that parameter with the command desc

```{r}
arrange(flights, desc(dep_delay))
```

Longest delay: 1301 minutes, or approx 21 hours

(Arranging is useful for finding smallest or largest values)

(usually, NA values put at end)

Can put in alphabetical order too: 

```{r}
arrange(flights, dest)
```

Or in reverse alphabetical order: 

```{r}
arrange(flights, desc(dest))
```


## Using select to pick out columns/factors

filter = choose a subset of rows based on criteria

select = choose a subset of columns/variables

```{r}
select(flights, year, dep_delay)
```

First argument = data set

Remaining arguments tell you which columns/variables to keep or lose, and the order to put them in. 

Keep all columns/variables in a range:

```{r}
select(flights, year:day)
```

Keep all columns except year: 

```{r}
select(flights, -year)
```

Put origin first, then dest, then everything else in its original order: 

```{r}
select(flights, origin, dest, everything())
```

On homework: put relevant columns (asked about in question) first, otherwise can't see them in knitted file


### Helper Functions

To pick out variable names that meet certain criteria:

starts_with()

```{r}
select(flights, starts_with("dep"))
```

contains()
```{r}
select(flights, contains("dep"))
```

ends_with()

```{r}
select(flights, ends_with("time"))
```

Can combine different ways of selecting columns/factors in one command: 
```{r}
flights_sml <- select(flights, 
                      year:day, 
                      ends_with("delay"), 
                      distance, 
                      air_time
                      )
flights_sml
```

R knows the command continues onto the next line because the open parentheses ( hasn't been closed yet


## Using mutate to create new columns

Example: in a spreadsheet, easy to make a new column that is the difference of two previous columns

For a tibble, do this with mutate()

Add two new columns:

```{r}
mutate(flights_sml, 
       gain = dep_delay - arr_delay, 
       speed = distance/air_time * 60 )
```

First argument = data set

Other arguments = new columns and how to compute them. 

Argument format: [new column name] = [formula for calculating it, in terms of other columns]

For Homework: then, put relevant columns first! 

```{r}
flights_sml_newcols <- mutate(flights_sml, 
       gain = dep_delay - arr_delay, 
       speed = distance / air_time * 60)
select(flights_sml_newcols, 
       dep_delay, arr_delay, 
       gain, distance, air_time, 
       speed, everything())
```

Check knitted file, including PDF, to be sure all show up! 

Also check: can see all of code (especially important when knitting directly to PDF)


### if/else function, pmin and pmax

Can also use if/else statements

For example, what if you want to know which of arr_delay and dep_delay is larger? 

ifelse function: 
1st argument is something that evaluates to TRUE or FALSE
2nd argument is output if first argument is TRUE
3rd argument is ourput if the first argument is FALSE

```{r}
flights_delay <- mutate(flights, 
                        larger_delay = ifelse(arr_delay >= dep_delay, 
                                              "arr", 
                                              "dep"))

select(flights_delay, dep_delay, arr_delay, larger_delay)
```


What if just want to put the larger of the two delay amounts in that column? 

```{r}
flights_delay <- mutate(flights, 
                        larger_delay = ifelse(arr_delay >= dep_delay, 
                                              arr_delay, 
                                              dep_delay))

select(flights_delay, dep_delay, arr_delay, larger_delay)
```

Are really just doing a max! 

**DON'T USE MAX() WITHIN MUTATE**

Is a max of two different columns, not a max of a single list, use pmax instead. 

```{r}
flights_pmax <- mutate(flights, larger_delay = pmax(arr_delay, dep_delay))

select(flights_pmax, dep_delay, arr_delay, larger_delay)
```

max() just finds the largest entry anywhere in both columns, and inserts that in new column: 

```{r}
flights_max <- mutate(flights, larger_delay = max(arr_delay, dep_delay))

select(flights_max, dep_delay, arr_delay, larger_delay)
```

(this is almost never what you want to do)


### str_c

What if you want to combine strings in different columns? Use str_c()

```{r}
mutate(flights, route = str_c(origin, " to ", dest))
```

Arguments can be columns in the data set or fixed strings (same string used for each row)


### rank()

Use to find the order of rows without actually reordering them

rank(list) returns the rank of each item in the list (from smallest to largest) 

```{r}
rank(c(9,5,8))
```

9 is the largest of the three numbers, so it has rank 3
5 is the smallest of the numbers, so it has rank 1
8 is the middle of the three numbers, so it has rank 2

```{r}
mutate(starwars, height_order = rank(height))
```

Luke Skywalker is the 28th shortest character in the dataset. 

Note: when there's a tie, can get fractional ranks

```{r}
rank(c(9,5,8,8))
```

Note: to rank in reverse order, use desc(), just like with arrange. 


# Filtering Helper Functions

Is there a version of starts_with(), ends_with(), contains() for filtering?

Yes, but it's tricky, and we'll do this later in the course. 

For contains(), use str_detect()

```{r}
filter(flights, str_detect(dest, "X"))
```

For more complicated things, use regular expressions; will spend a whole week on this!

Additional helper functions for filter(): Wednesday's preclass video


# Adding TRUE, FALSE

In R, TRUE values are treated as 1s, FALSE values are treated as 0s

```{r}
x <- c(TRUE, FALSE, TRUE, TRUE)

sum(x) # Does 1 + 0 + 1 + 1 = 3
```

When you sum a bunch of TRUE/FALSE values, the answer is the number of TRUEs

This can be very useful for understanding data

How many flights have dep_delay > 60? 

```{r}
sum( flights$dep_delay > 60 ) 
# You get NA because when you sum thins and one element is NA< the whole sum is NA
```
```{r}
sum( is.na(flights$dep_delay) )
```
There are 8255 flights that have a departure delay of NA. 


# Summary

Can transform data via: 

filter: pick out a subset of rows

arrange: sort rows

select: pick out a subset of columns

mutate: create new columns

For all, first input is the data frame, remaining inputs say what to do with it. 

When you sum a bunch of TRUE/FALSE values, the answer is the number of TRUEs


# Activity 

On Canvas, Files > Class Material > 9-9 Activity

If finish all regular questions, move on to bonus question.

If finish bonus question, can leave: check in with Prof. Cannon before leaving

Feel free to talk to/work with others around you.

