---
title: "8-26 Class Notes"
author: "Sarah Cannon"
date: "2024-08-26"
output: html_document
---


# Slides

Can be found on Canvas, in the Files section


# Overview of R, RStudio

R - programming language we'll be using this semester

RStudio - the development environment (program) we'll be using this semester to write code in R

Can run commands in console (bottom left by default), such as: 

2+3 
  (outputs 5)
  
x <- 3
  (stores the value 3 in a variable x, no output)
  
y <- 2 
  (stores the value 2 in variable y, no output)
  
x
  (outputs 3)
  
y 
  (outputs 2)
  
x + y 
  (outputs 5)
  
x - y
  (outputs 1)
  
Notes: 
- no output when storing a variable
- can see current variables in the Environment panel (top right) 


# RMarkdown 

However, we'll rarely use the console (except to install packages)

Instead, we'll use RMarkdown

File > New > RMarkdown

Or, simply open a .rmd file 
  (You can find these notes as "8-26 Class Notes.rmd" in the Files section of Canvas)

RMarkdown: 

- A way of organizing code into a single file that you can run in chunks or run all at once (Similarish to Jupyter Notebooks)
- File > New File > RMarkdown. Files end in .rmd
- For now, delete default text after header
- Can insert code chunks in RMarkdown, and run commands there just like you would in the console by clicking green arrow 

```{r}
3+2
```
(Shortcut to insert code chunk: CTRL+ALT+I or CMD+Option+I)

- Can run multiple commands at once:

```{r}
x <- 3
y <- 2
x+y
x-y
```

# Data and Plots

Just enough to get you started! 

Look at built-in data set trees: 
```{r}
trees 
```

Get more information about the data set trees: 
```{r}
?trees
```

Make a scatterplot with Girth and Height: 
```{r}
plot(trees$Girth, trees$Height)
```


# Knitting

Produces a nice version of your document, with both text and code! We'll talk more about knitting in the second class.


# Activity

Classes will end with an activity!  Find it on Canvas, in the Files section ("8-26 Activity.pdf")


