---
title: "10-28 Activity Solutions"
author: "Sarah Cannon"
date: "10/28/2023"
output: html_document
---

```{r}
library(DBI)
library(RSQLite)
```

> 1. **Connect to the database Northwind_small.sqlite, which is on canvas. What tables are in this database?** 




>2. **Change the start of this code chunk so that this SQL command runs correctly. How many employees does this company have? **

```{r}
SELECT * FROM Employee
```

> 3. **Make a smaller table containing only Employee ID, First and Last Name, and City**



> 4. **How many different values are there in the TitleOfCourtesy column? Use code to find out; your command should still work even if the Employee table were to have many more rows than it currently does.**



> 5. **Filter the entire Employees table to only keep those Employees where the word "degree" appears in the Notes column. Note: there should be four.** 




> 6. **Arrange the employees in the Employee table by LastName.  Which employee comes first?**




> 7. **Disconnect from the database.**




> BONUS A: **There's a second way to use RStudio to access remote databases, using the dbSendQuery and dbFetch commands. Here's an example of the same code from Question 2:**

```{r}
df7 <- dbSendQuery(db_con, "SELECT * FROM Employee")
dbFetch(df7)
```

>**Produce the same output as in Question 4, but using an R code block rather than an SQL code block.**





> BONUS B: **In the tidyverse, we used the functions unite or mutate & str_c to combine two columns.  In SQL, this is done by a single keyword, CONCAT, which takes as arguments the things to combine, such as CONCAT(Col1, Col2, "!"), which combines columns 1 and 2, followed by an exclamation mark. Combine the FirstName and LastName for all the Employees into a single column called Name; you'll have to use other keyword(s) in addition to just CONCAT.**





