---
title: "9-9 In-class Activity"
author: "Sarah Cannon"
date: "9/9/2024"
output:
  html_document: default
  pdf_document: default
---


```{r, message = FALSE}
library(tidyverse)
```

We'll be using the starwars data set.  First, just take a look at it and familiarize yourself with it: 

```{r}
starwars
```


>1. **Give a command that outputs the name, gender, homeworld, and birth year variables from starwars ** 

```{r}
starwars %>%
  select(name, gender, homeworld, birth_year)

```


>2. **What three characters have the smallest masses? Write a command that will show them. **

```{r}
starwars %>%
  arrange(mass) %>%
  select(name, mass) %>%
  filter(!is.na(mass)) %>%
  head(3)
```


>3. **Give a command that adds a column to the data set called mass_height_ratio, which is a character's mass divided by their height.** 

```{r}
starwars %>%
  mutate(mass_height_ratio = mass / height)
```



>4. **Give a command that returns a table containing all characters that have mass greater than 136 or have blue eyes. How many are there?** 

```{r}
result <- starwars %>%
  filter(mass > 136 | eye_color == "blue")

result

nrow(result)
```


>5. **Give a command that returns all characters whose mass is not available.** 

```{r}
starwars %>%
  filter(is.na(mass))
```


>6. **Give a command or series of commands that reorders the variables/columns of starwars, so that all the variables/columns with an "e" in the variable/column name appear first, followed by all the other variables/columns in their original order.**

```{r}
cols_with_e <- names(starwars)[grepl("e", names(starwars))]
cols_without_e <- setdiff(names(starwars), cols_with_e)

starwars %>%
  select(all_of(cols_with_e), all_of(cols_without_e))
```



>BONUS 1: **Give a single command that return a data frame that contains all characters with blonde hair.  Hint: there are 4. ** 

```{r}
starwars %>%
  filter(str_detect(str_to_lower(hair_color), "blond"))

```



>BONUS 2: **Make a data table with the variables name, hair_color, skin_color, eye_color, and weight in pounds (not kilograms). It should contain all human characters, and be sorted by their birth year.** 

```{r}
starwars %>%
  filter(species == "Human") %>%
  mutate(weight_lbs = mass * 2.20462) %>%
  select(name, hair_color, skin_color, eye_color, weight_lbs, birth_year) %>%
  arrange(birth_year)
```







